---
title: "PHASE 1 - TopGO Directional (up/down) Regulation"
author: "Megan Delatte-Guidry"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Gene Ontology Enrichment Analysis on DEG gene sets - UP and DOWN regulated genes as separate datasets for each phase!

```{r}
library(topGO)
library(stringr)
library(dplyr)
library(tidyr)
```

### Steps

1. Data preparation: List of genes identifiers, gene scores, list of differentially expressed genes or a criteria for selecting genes based on their scores, as well as gene-to-GO annotations are all collected and stored in a single R object.   
(referenced [Amy Zyck's TopGO script](https://github.com/MarineEvoEcoLab/ROD_CADO/blob/main/Outlier_Exploration/Outlier_GO.Rmd) for the gene2go table)

2. Running the enrichment tests: Using the object created in the first step the user can perform enrichment analysis using any feasible mixture of statistical tests and methods that deal with the GO topology.

3. Analysis of the results: The results obtained in the second step are analysed using summary functions and visualization tools.
_________________________________________________________________________________

## Step 1: Data preparation 

### gene2go table from annotation information

PULLING gene2go_full.tab FROM THE WGCNA DIRECTORY (../wgcna/gene2go_full.tab)

gaf from NCBI with the GO terms for genes in the eastern oyster genome. I'm going to take this file and extract the Gene IDs and associated GO terms.

In terminal:
```{bash}
#wget https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/002/022/765/GCF_002022765.2_C_virginica-3.0/GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf.gz
#gunzip GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf.gz
#mv GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf gene_ontology.gaf
```

Now I can make a new .tab file with just the gene ID and GO IDs (3rd and 5th columns) renamed "gene_id" and "GO_Annotation". The gene_name column will have one unique observation per row and all associated GO terms will be saved in the same row, with values separated by ;

```{bash}
#awk 'NR>9 {print $3 "\t" $5}' gene_ontology.gaf | 
#sed '1i gene_id\tGO_Annotation' | 
#sort | awk '
#NR==1 {print; next}
#$1!=prev {
#if (NR>2) print prev "\t" goids;
#prev=$1;
#goids=$2;
#next
#}
#{
#goids=goids "; " $2
#}
#END {print prev "\t" goids}
#' > gene2go_full.tab
```

Gene IDs for down and upregulated genes are in the following files:
outputs/phase2_stress_directional_expression.csv

Make separate files with gene IDs and GO IDs for UP & DOWN regulated genes
```{r}
# load data
p2_stress <- read.csv("outputs/phase2_stress_directional_expression.csv", header = TRUE, sep = ",")
p2_stress
# separate UP and DOWN reg genes into two data sets
p2_stress_up <- subset(p2_stress, diffexpressed=="UP")
p2_stress_down <- subset(p2_stress, diffexpressed=="DOWN")

# load go annotation file to assign go terms to color module
go_terms <- read.table("../wgcna/gene2go_full.tab", header = TRUE, sep="\t")
head(go_terms)

# UP - reformat gene_id column to be just LOC#, add in go annotation, remove 'MSTRG' rows and NA rows
p2_stress_up$gene_id <- str_extract(p2_stress_up$GeneID, "[^|]+$")
dim(p2_stress_up)
p2_stress_up_wGO <- merge(p2_stress_up, go_terms, by = "gene_id", all.x = TRUE)
dim(p2_stress_up_wGO)
#clean up
p2_stress_up_wGO <- na.omit(p2_stress_up_wGO)
p2_stress_up_wGO <- p2_stress_up_wGO[, c("gene_id", "GO_Annotation")]
dim(p2_stress_up_wGO)

# DOWN - reformat gene_id column to be just LOC#, add in go annotation, remove 'MSTRG' rows and NA rows
p2_stress_down$gene_id <- str_extract(p2_stress_down$GeneID, "[^|]+$")
dim(p2_stress_down)
p2_stress_down_wGO <- merge(p2_stress_down, go_terms, by = "gene_id", all.x = TRUE)
dim(p2_stress_down_wGO)
#clean up
p2_stress_down_wGO <- na.omit(p2_stress_down_wGO)
p2_stress_down_wGO <- p2_stress_down_wGO[, c("gene_id", "GO_Annotation")]
dim(p2_stress_down_wGO)

# write out tables
#write.table(p2_stress_up_wGO, "p2_stress_UP_genes_GO_header.tab", sep = "\t", row.names=FALSE)
#write.table(p2_stress_down_wGO, "p2_stress_DOWN_genes_GO_header.tab", sep = "\t", row.names=FALSE)
```

Read in expressed genes file - maybe I'll use this??
```{r}
p2_filt_counts <- read.csv("../de_analysis/outputs/p2_filtered_gene_count_matrix.csv")
head(p2_filt_counts)
```


Read in gene2go information 

```{r}
all_gene2go<-read.delim("../wgcna/gene2go_full.tab", sep="\t")
UP_gene2go<-read.delim("p2_stress_UP_genes_GO_header.tab", sep="\t")
DOWN_gene2go<-read.delim("p2_stress_DOWN_genes_GO_header.tab", sep="\t")
View(UP_gene2go)
View(DOWN_gene2go)
```

Make list of genes for input to TopGO
```{r}
#UP_genes <- as.character(unique(UP_gene2go$gene_id))
DOWN_genes <- as.character(unique(DOWN_gene2go$gene_id))

all_genes <- as.character(unique(all_gene2go$gene_id))

#UP_GeneList <- factor(as.integer(all_genes %in% UP_genes))
#names(UP_GeneList) <- all_genes

DOWN_GeneList <- factor(as.integer(all_genes %in% DOWN_genes))
names(DOWN_GeneList) <- all_genes
```

The following code will perform GO enrichment using the weighted Fisher's exact test  

Read in gene-to-go-mappings
```{r}
gene2go_topgo <-readMappings("../wgcna/gene2go_full.tab", IDsep=";", sep="\t")
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```

### Biological Processes

Create `topGOdata` object, which is required for topGO analysis
```{r}
#GO_UP_BP <-new("topGOdata", ontology="BP", gene2GO=gene2go_topgo, allGenes=UP_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)

GO_DOWN_BP <-new("topGOdata", ontology="BP", gene2GO=gene2go_topgo, allGenes=DOWN_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
#GO_UP_BP_FE <- runTest(GO_UP_BP, algorithm="weight01", statistic="fisher")

GO_DOWN_BP_FE <- runTest(GO_DOWN_BP, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
#GO_UP_BP_RT <- GenTable(GO_UP_BP, Fisher = GO_UP_BP_FE, orderBy = "Fisher",  topNodes = 50, numChar = 100)

GO_DOWN_BP_RT <- GenTable(GO_DOWN_BP, Fisher = GO_DOWN_BP_FE, orderBy = "Fisher",  topNodes = 50, numChar = 100)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
#GO_UP_BP_RT$Fisher<-as.numeric(GO_UP_BP_RT$Fisher)
#GO_UP_BP_RT_sig<-GO_UP_BP_RT[GO_UP_BP_RT$Fisher<0.05,]

GO_DOWN_BP_RT$Fisher<-as.numeric(GO_DOWN_BP_RT$Fisher)
GO_DOWN_BP_RT_sig<-GO_DOWN_BP_RT[GO_DOWN_BP_RT$Fisher<0.05,]
```

Merge `GO_DOWN_BP_RT_sig` with GO and gene info. 
```{r}
# Separate GO terms 
DOWN_gene2go <- DOWN_gene2go %>%
  separate_rows(GO_Annotation, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
DOWN_gene2go$GO_Annotation <- trimws(DOWN_gene2go$GO_Annotation)
GO_DOWN_BP_RT_sig$GO.ID <- trimws(GO_DOWN_BP_RT_sig$GO.ID)

# Join the datasets based on GO term
GO_DOWN_BP_RT_sig_gene <- DOWN_gene2go %>%
  left_join(GO_DOWN_BP_RT_sig, by = c("GO_Annotation" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_DOWN_BP_RT_sig_gene$ontology <- "Biological Processes"
```

### Cellular Components
Create `topGOdata` object, which is required for topGO analysis
```{r}
#GO_UP_CC <-new("topGOdata", ontology="CC", gene2GO=gene2go_topgo, allGenes=UP_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)

GO_DOWN_CC <-new("topGOdata", ontology="CC", gene2GO=gene2go_topgo, allGenes=DOWN_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
#GO_UP_CC_FE <- runTest(GO_UP_CC, algorithm="weight01", statistic="fisher")

GO_DOWN_CC_FE <- runTest(GO_DOWN_CC, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
#GO_UP_CC_RT <- GenTable(GO_UP_CC, Fisher = GO_UP_CC_FE, orderBy = "Fisher",  topNodes = 50, numChar = 100)

GO_DOWN_CC_RT <- GenTable(GO_DOWN_CC, Fisher = GO_DOWN_CC_FE, orderBy = "Fisher",  topNodes = 50, numChar = 100)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
#GO_UP_CC_RT$Fisher<-as.numeric(GO_UP_CC_RT$Fisher)
#GO_UP_CC_RT_sig<-GO_UP_CC_RT[GO_UP_CC_RT$Fisher<0.05,]

GO_DOWN_CC_RT$Fisher<-as.numeric(GO_DOWN_CC_RT$Fisher)
GO_DOWN_CC_RT_sig<-GO_DOWN_CC_RT[GO_DOWN_CC_RT$Fisher<0.05,]
```

Merge `GO_DOWN_CC_RT_sig` with GO and gene info. 
```{r}
# Separate GO terms 
DOWN_gene2go <- DOWN_gene2go %>%
  separate_rows(GO_Annotation, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
DOWN_gene2go$GO_Annotation <- trimws(DOWN_gene2go$GO_Annotation)
GO_DOWN_CC_RT_sig$GO.ID <- trimws(GO_DOWN_CC_RT_sig$GO.ID)

# Join the datasets based on GO term
GO_DOWN_CC_RT_sig_gene <- DOWN_gene2go %>%
  left_join(GO_DOWN_CC_RT_sig, by = c("GO_Annotation" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_DOWN_CC_RT_sig_gene$ontology <- "Cellular Components"
```


### Molecular Functions

Create `topGOdata` object, which is required for topGO analysis
```{r}
#GO_UP_MF <-new("topGOdata", ontology="MF", gene2GO=gene2go_topgo, allGenes=UP_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)

GO_DOWN_MF <-new("topGOdata", ontology="MF", gene2GO=gene2go_topgo, allGenes=DOWN_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
#GO_UP_MF_FE <- runTest(GO_UP_MF, algorithm="weight01", statistic="fisher")

GO_DOWN_MF_FE <- runTest(GO_DOWN_MF, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
#GO_UP_MF_RT <- GenTable(GO_UP_MF, Fisher = GO_UP_MF_FE, orderBy = "Fisher",  topNodes = 50, numChar = 100)

GO_DOWN_MF_RT <- GenTable(GO_DOWN_MF, Fisher = GO_DOWN_MF_FE, orderBy = "Fisher",  topNodes = 50, numChar = 100)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
#GO_UP_MF_RT$Fisher<-as.numeric(GO_UP_MF_RT$Fisher)
#GO_UP_MF_RT_sig<-GO_UP_MF_RT[GO_UP_MF_RT$Fisher<0.05,]

GO_DOWN_MF_RT$Fisher<-as.numeric(GO_DOWN_MF_RT$Fisher)
GO_DOWN_MF_RT_sig<-GO_DOWN_MF_RT[GO_DOWN_MF_RT$Fisher<0.05,]
```

Merge `GO_DOWN_MF_RT_sig` with GO and gene info. 
```{r}
# Separate GO terms 
DOWN_gene2go <- DOWN_gene2go %>%
  separate_rows(GO_Annotation, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
DOWN_gene2go$GO_Annotation <- trimws(DOWN_gene2go$GO_Annotation)
GO_DOWN_MF_RT_sig$GO.ID <- trimws(GO_DOWN_MF_RT_sig$GO.ID)

# Join the datasets based on GO term
GO_DOWN_MF_RT_sig_gene <- DOWN_gene2go %>%
  left_join(GO_DOWN_MF_RT_sig, by = c("GO_Annotation" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_DOWN_MF_RT_sig_gene$ontology <- "Molecular Functions"
```

### Join ontologies

Bind `GO_UP_BP_RT_sig_gene`, `GO_UP_CC_RT_sig_gene`, and `GO_UP_MF_RT_sig_gene` to make a df that has significantly enriched GO terms for all ontologies 
```{r}
#GO_UP_sig_genes <- rbind(GO_UP_BP_RT_sig_gene, GO_UP_CC_RT_sig_gene, GO_UP_MF_RT_sig_gene)
GO_DOWN_sig_genes <- rbind(GO_DOWN_BP_RT_sig_gene, GO_DOWN_CC_RT_sig_gene, GO_DOWN_MF_RT_sig_gene)
```

Merge `pi_bed_filt` with `GO_UP_sig_genes` and calculate proportion of significant genes
```{r}
merged_down_genes <- GO_DOWN_sig_genes %>%
  mutate(prop.sig.genes = Significant/Annotated)

# saving unique instances of gene/GO pairs
merged_down_genes <- unique(merged_down_genes)
```

```{r}
# Write to csv 
#write.csv(merged_down_genes, "p2_stress_DOWN_genes_GO_enrichment.csv")
```



Plot - need to work out 
```{r}
GO_plot_pi <- ggplot(merged_out_genes, aes(x = Term, y = -log(Fisher), size = prop.sig.genes, fill = -log(Fisher))) +
  expand_limits(y = 1.5) +
  ylim(1, 7.25) +
  #Add horizontal lines with a single aesthetic value
  geom_hline(yintercept = -log10(0.01), linetype = "longdash", colour = "black", linewidth = .6) +
  geom_hline(yintercept = -log10(0.001), linetype = "solid", colour = "black", linewidth = .6) +
  geom_point(shape = 21) + 
  scale_size(range = c(2, 12)) + 
  scale_fill_continuous(low = "#1AD3D1FF", high = "#4686FBFF") +
  xlab('') + 
  ylab('Enrichment score') +
  #labs(caption = 'Cut-off lines at p=0.01 and p=0.001') +
  theme_bw() +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); GO_plot_pi

ggsave(
  "GO_plot_pi.png",
  plot = GO_plot_pi,
  width = 8,
  height = 10,
  dpi = 300
)

```

Lastly, I'm just going to save the figure as a .png
```{r}
# Open PNG device with high resolution and transparent background
png(filename = "GO_plot_pi.png",
    width = 2000,   # width in pixels (adjust for desired size)
    height = 2000,  # height in pixels
    res = 200,      # resolution in dpi for high quality
    bg = "transparent")  # transparent background

# Print the plot object
print(GO_plot_pi)

# Close the device to finalize the file
dev.off()
```