---
title: "Ch 3 Volcano Plots"
author: "Megan Delatte-Guidry"
date: "`r Sys.Date()`"
output: html_document
---

A volcano plot is a of scatterplot that shows statistical significance (p-value) versus magnitude of change (fold change). It is a great way of visualising the results from differential gene expression analysis.

After visualizing the directionality and significance of expressed genes in these volcano plots, I ran a TopGO analysis on the datasets of those identified genes. 

source for plots: https://biostatsquid.com/volcano-plots-r-tutorial/

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Loading relevant libraries 
#library(tidyverse) # includes ggplot2, for data visualization. dplyr, for data manipulation.
library(ggplot2)
library(dplyr)
library(RColorBrewer) # for a colourful plot
library(ggrepel) # for nice annotations
library(stringr)
```

## PHASE I 

Load in and clean up data
```{r}
# Import DEG results - full data set with sig and nonsig 
deg1 <- read.csv("../de_analysis/outputs/phase1_res_stress.csv")

# make a new column called "GeneID" in the 'deg' df that is only the gene ID
deg1$GeneID <- str_extract(deg1$X, "[^|]+$")

#clean up data frame
deg1_clean <- deg1[,c(8,3,6,7)]
colnames(deg1_clean)
```


Add a column to the data frame to specify if they are UP- or DOWN- regulated (log2fc respectively positive or negative)
```{r}
# create new column and fill values with "NO" as default
deg1_clean$diffexpressed <- "NO"
colnames(deg1_clean)

# set "UP" based on previous threshold
deg1_clean$diffexpressed[deg1_clean$log2FoldChange > 0.8 & deg1_clean$padj < 0.001] <- "UP"
# set "DOWN" based on previous threshold
deg1_clean$diffexpressed[deg1_clean$log2FoldChange < -0.8 & deg1_clean$padj < 0.001] <- "DOWN"
# Explore a bit
head(deg1_clean[order(deg1_clean$padj) & deg1_clean$diffexpressed == 'DOWN', ])

# variable with only up and down
deg1_clean_direction <- subset(deg1_clean, diffexpressed != 'NO')

#check we got all of the data - should return "TRUE"
all(nrow(deg1_clean_direction) == sum(nrow(deg1_clean[deg1_clean$diffexpressed == 'DOWN', ]) + nrow(deg1_clean[deg1_clean$diffexpressed == 'UP', ])))

# write out dataset for topGO
#write.csv(deg1_clean_direction, "outputs/phase1_directional_expression.csv")
```



Plotting
Note: consider adding GO annotations to plot
```{r}
# Create a basic volcano plot with adjusted p-values (to account for multiple hypothesis testing)
ggplot(data = deg1_clean, aes(x = log2FoldChange, y = -log10(padj), fill = diffexpressed))  +
  #add threshold lines (these are mostly arbitrarily chosen)
  geom_vline(xintercept = c(-0.8, 0.8), col = "gray", linetype = 'dashed') +
  #chose a pretty stringent p-value cut off 
  geom_hline(yintercept = -log10(0.001), col = "gray", linetype = 'dashed') +
  geom_point(pch=21, size = 3) +
  annotate("text", x = -10, y = 10, label = "Downregulated", color = "cornflowerblue", size = 6) +
  annotate("text", x = 10, y = 10, label = "Upregulated", color = "maroon", size = 6) +
  # to set the colours of our variable
  scale_fill_manual(values = c("cornflowerblue", "grey", "maroon")) 
                     # to set the labels in case we want to overwrite the categories from the dataframe (UP, DOWN, NO)
                     #labels = c("Downregulated", "Not significant", "Upregulated")) #removed legend
  # Biostatsquid theme
  theme_set(theme_classic(base_size = 20) +
              theme(
              axis.title.y = element_text(face = "bold",
                                          margin = margin(0,20,0,0),
                                          size = rel(1.1),
                                          color = 'black'),
              axis.title.x = element_text(hjust = 0.5,
                                          face = "bold",
                                          margin = margin(20,0,0,0),
                                          size = rel(1.1),
                                          color = 'black'),
              plot.title = element_text(hjust = 0.5),
              legend.position = "none"
              #legend.title = element_blank()
            ))
ggsave("plots/phase1_volcano.png", last_plot())
```


## PHASE II

Load in and clean up data
```{r}
# Import DEG results - full data set with sig and nonsig 
deg2_str <- read.csv("../de_analysis/outputs/phase2_res_stress.csv")
deg2_rec <- read.csv("../de_analysis/outputs/phase2_res_recovery.csv")
deg2_nai <- read.csv("../de_analysis/outputs/phase2_res_naive.csv")

# make a new column called "GeneID" in the 'deg' df that is only the gene ID
deg2_str$GeneID <- str_extract(deg2_str$X, "[^|]+$")
deg2_rec$GeneID <- str_extract(deg2_rec$X, "[^|]+$")
deg2_nai$GeneID <- str_extract(deg2_nai$X, "[^|]+$")

#clean up data frame
deg2_str_clean <- deg2_str[,c(8,3,6,7)]; colnames(deg2_str_clean)
deg2_rec_clean <- deg2_rec[,c(8,3,6,7)]; colnames(deg2_rec_clean)
deg2_nai_clean <- deg2_nai[,c(8,3,6,7)]; colnames(deg2_nai_clean)
```


Add a column to the data frame to specify if they are UP- or DOWN- regulated (log2fc respectively positive or negative)
```{r}
# create new column and fill values with "NO" as default
deg2_str_clean$diffexpressed <- "NO" ; colnames(deg2_str_clean)
deg2_rec_clean$diffexpressed <- "NO" ; colnames(deg2_rec_clean)
deg2_nai_clean$diffexpressed <- "NO" ; colnames(deg2_nai_clean)

# set "UP" based on previous threshold
deg2_str_clean$diffexpressed[deg2_str_clean$log2FoldChange > 0.8 & deg2_str_clean$padj < 0.001] <- "UP"
deg2_rec_clean$diffexpressed[deg2_rec_clean$log2FoldChange > 0.8 & deg2_rec_clean$padj < 0.001] <- "UP"
deg2_nai_clean$diffexpressed[deg2_nai_clean$log2FoldChange > 0.8 & deg2_nai_clean$padj < 0.001] <- "UP"
# set "DOWN" based on previous threshold
deg2_str_clean$diffexpressed[deg2_str_clean$log2FoldChange < -0.8 & deg2_str_clean$padj < 0.001] <- "DOWN"
deg2_rec_clean$diffexpressed[deg2_rec_clean$log2FoldChange < -0.8 & deg2_rec_clean$padj < 0.001] <- "DOWN"
deg2_nai_clean$diffexpressed[deg2_nai_clean$log2FoldChange < -0.8 & deg2_nai_clean$padj < 0.001] <- "DOWN"
# Explore a bit
head(deg2_str_clean[order(deg2_str_clean$padj) & deg2_str_clean$diffexpressed == 'DOWN', ])
head(deg2_rec_clean[order(deg2_rec_clean$padj) & deg2_rec_clean$diffexpressed == 'DOWN', ])
head(deg2_nai_clean[order(deg2_nai_clean$padj) & deg2_nai_clean$diffexpressed == 'DOWN', ])

# variable with only up and down
deg2_str_clean_direction <- subset(deg2_str_clean, diffexpressed != 'NO')
deg2_rec_clean_direction <- subset(deg2_rec_clean, diffexpressed != 'NO')
deg2_nai_clean_direction <- subset(deg2_nai_clean, diffexpressed != 'NO')

#check we got all of the data - should return "TRUE"
all(nrow(deg2_str_clean_direction) == sum(nrow(deg2_str_clean[deg2_str_clean$diffexpressed == 'DOWN', ]) + nrow(deg2_str_clean[deg2_str_clean$diffexpressed == 'UP', ])))
all(nrow(deg2_rec_clean_direction) == sum(nrow(deg2_rec_clean[deg2_rec_clean$diffexpressed == 'DOWN', ]) + nrow(deg2_rec_clean[deg2_rec_clean$diffexpressed == 'UP', ])))
all(nrow(deg2_nai_clean_direction) == sum(nrow(deg2_nai_clean[deg2_nai_clean$diffexpressed == 'DOWN', ]) + nrow(deg2_nai_clean[deg2_nai_clean$diffexpressed == 'UP', ])))

# write out dataset for topGO
#write.csv(deg2_str_clean_direction, "outputs/phase2_stress_directional_expression.csv")
#write.csv(deg2_rec_clean_direction, "outputs/phase2_recovery_directional_expression.csv")
#write.csv(deg2_nai_clean_direction, "outputs/phase2_naive_directional_expression.csv")
```



Plotting
Note: consider adding GO annotations to plot
```{r}
# Create a basic volcano plot with adjusted p-values (to account for multiple hypothesis testing)
ggplot(data = deg2_str_clean, aes(x = log2FoldChange, y = -log10(padj), fill = diffexpressed))  +
  #add threshold lines (these are mostly arbitrarily chosen)
  geom_vline(xintercept = c(-0.8, 0.8), col = "gray", linetype = 'dashed') +
  #chose a pretty stringent p-value cut off 
  geom_hline(yintercept = -log10(0.001), col = "gray", linetype = 'dashed') +
  geom_point(pch=21, size = 3) +
  annotate("text", x = -10, y = 10, label = "Downregulated", color = "cornflowerblue", size = 6) +
  annotate("text", x = 10, y = 10, label = "Upregulated", color = "maroon", size = 6) +
  annotate("text", x= -20, y= 28, label = "Stress", size = 10) +
  # to set the colours of our variable
  scale_fill_manual(values = c("cornflowerblue", "grey", "maroon")) 
                     # to set the labels in case we want to overwrite the categories from the dataframe (UP, DOWN, NO)
                     #labels = c("Downregulated", "Not significant", "Upregulated")) #removed legend
  # Biostatsquid theme
  theme_set(theme_classic(base_size = 20) +
              theme(
              axis.title.y = element_text(face = "bold",
                                          margin = margin(0,20,0,0),
                                          size = rel(1.1),
                                          color = 'black'),
              axis.title.x = element_text(hjust = 0.5,
                                          face = "bold",
                                          margin = margin(20,0,0,0),
                                          size = rel(1.1),
                                          color = 'black'),
              plot.title = element_text(hjust = 0.5),
              legend.position = "none"
              #legend.title = element_blank()
            ))
ggsave("plots/phase2_stress_volcano.png", last_plot())
```
