---
title: "GO_MWU"
author: "Megan Delatte-Guidry"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Rank-based Gene Ontology analysis of gene expression data with [`GO_MWU`](https://github.com/z0on/GO_MWU) 

##Short Guide (pulled from package github):
1. Put all this into the same directory:
  * scripts: GO_MWU.R, gomwu_a.pl, gomwu_b.pl, gomwu.functions.R
  * GO hierarchy file (go.obo, http://www.geneontology.org/GO.downloads.ontology.shtml)
  * table of GO annotations for your sequences: two-column (gene id - GO terms), tab-delimited, one line per gene, multiple GO terms separated by semicolon. If you have multiple lines per gene, use nrify_GOtable.pl to merge them. Do NOT include genes without GO annotations.
  * table of measure of interest for your sequences: two columns of comma-separated values: gene id, continuous measure of change such as                 log(fold-change). To perform standard GO enrichment analysis based on Fisher's exact test, use binary measure (1 or 0, i.e., either sgnificant or not). To analyze modules derived from WGCNA, specify 0 for genes not included in the module and the kME value (number between 0 and 1, module membership score) for genes included in the module.

It is important to have the latter two tables representing the whole genome (or transcriptome) - at least the portion that was measured - rather than some select group of genes since the test relies on comparing the behavior of individual GO categories to the whole.

## Workflow attempt #1 
- generate full DEG table (for each phase) w/o removing n.s. data with de_analysis/DESeq2_analysis.Rmd
- assign GO terms with de_analysis/GO_term_assignment.Rmd script
- reformat GO term table and LFC table (see steps below)
- remove duplicates with the nrify_GOtable.pl

### Files needed
DEG table generated with DESeq2 with assigned GO terms  
*reformatted into a GO term table and a LFC table below*

**Formatting GO annotation file**
```{bash}
cd table_formatting
cp ../../de_analysis/outputs/deg_phase1_wGO.csv .
#convert to tab delimited
sed 's/,/\t/g' deg_phase1_wGO.csv > deg_phase1_wGO.tsv
#remove quotes
#sed 's/"//g' deg_phase1_wGO.tsv > deg_phase1_forGOMWU.tsv
```

**More formatting - easier for me to do in R**
```{r}
#library(dplyr)
#read in file (setting empty strings and NAs as "NA")
#table <- read.table("table_formatting/deg_phase1_wGO.tsv", sep="\t", header=TRUE, na.strings = c("", "NA"))
#table <- read.table("table_formatting/deg_phase1_wGO.tsv", sep="\t", header=TRUE)
#pull out only GeneID column and Annotation_GO_ID column
#table2 <- table[, c("GeneID","Annotation_GO_ID")]
#clean up the table (remove rows with NAs)
#table_clean <- na.omit(table2)
#write out table
#write.table(table_clean, "phase1_GOtable.tsv", sep="\t", row.names = FALSE, quote = FALSE)

#load packages
library(dplyr)
# read in file
table <- read.table("table_formatting/deg_phase1_wGO.tsv",
                    sep = "\t", header = TRUE, stringsAsFactors = FALSE)
#pull out only GeneID column and Annotation_GO_ID column
table2 <- table %>% select(GeneID, Annotation_GO_ID)
#clean up table - reassign blank or NA GO annotations as "unknown"
table_clean <- table2 %>%
  mutate(Annotation_GO_ID = as.character(Annotation_GO_ID)) %>%
  mutate(Annotation_GO_ID = ifelse(
    is.na(Annotation_GO_ID) | Annotation_GO_ID == "",
    "unknown",
    Annotation_GO_ID
  ))
#write out table with out vector numbers and quotations
write.table(table_clean, "table_formatting/phase1_GOtable.tsv",
            sep = "\t", row.names = FALSE, quote = FALSE)
```


**Formatting LFC table**
```{r}
#read in file (setting empty strings and NAs as "NA")
lfc <- read.table("table_formatting/deg_phase1_wGO.tsv", sep="\t", header=TRUE, na.strings = c("", "NA"))
#pull out only GeneID column and Annotation_GO_ID column
lfc2 <- lfc[, c("GeneID","log2FoldChange")]
#clean up the table (remove rows with NAs) - omitting this step bc these were changed to 'unknown' in GO table not removed
#lfc_clean <- na.omit(lfc2)
#write out table
write.csv(lfc2, "table_formatting/phase1_lfc_table.csv", row.names = FALSE, quote= FALSE)
```

Check for repeated lines for the same gene in data! GO MWU doesn't like duplicates, so we'll run the nrify_GOtable.pl script to deal with duplicates before proceeding with the GO_MWU.R script.  
nrify_GOtable.pl will merge the repeated lines. 
```{r}
#check for duplicates in GO table
sum(duplicated(table_clean$GeneID))

#check for duplicates in LFC table
sum(duplicated(lfc2$GeneID))
```

Remove duplicates in GeneID column for GO table. Don't think I have to remove the duplicates in the LFC table?

"The table must contain just a single line per gene. If you have a table in which a gene appears on several lines, use the included perl script nrify_GOtable.pl to compile a non-redundant table. nrify_GOtable.pl takes a single argument, which is the filename to process, and prints the result to STDOUT, so use it like this: nrify_GOtable.pl [filename of the redundant table] > [filename of the non-redundant table]"

```
perl nrify_GOtable.pl table_formatting/phase1_GOtable.tsv > dup_phase1_GOtable.tsv
#perl nrify_GOtable.pl table_formatting/phase1_lfc_table.csv > dup_phase1_lfc_table.csv
```

Check
```{r}
dup_GO <- read.table("dup_phase1_GOtable.tsv")
#check for duplicates in GO table
sum(duplicated(dup_GO$GeneID))

#dup_lfc <- read.csv("dup_phase1_lfc_table.csv")
#check for duplicates in LFC table
#sum(duplicated(dup_lfc$GeneID))

dim(dup_GO)
dim(lfc2)
```


2. Make sure you have perl and R. Windows-based people: install perl from here and specify your perl.exe file with full path to it as the perlPath argument to function gomwuStats (inside the GO_MWU.R script). The R part requires package "ape", which you might need to install prior to running this method.

3. Open GO_MWU.R script; edit the input file names, mark and execute bits of code separated by blank lines one by one. Follow instructions given as comments in the script.

4. Drag corner of the plot to rescale and match text and tree. After this, to achieve better "word map" effect, rerun gomwuPlot with modified "txtsize" parameter.

5. Save the plot as pdf file.



## Workflow attempt #2 
- generate full DEG table (for each phase) w/o removing n.s. data with de_analysis/DESeq2_analysis.Rmd
- reformat GeneID column of DEG table
- calculate signed negative log p-value and output table with gene id and signed neg log p-value
- remove duplicates from DEG table with the nrify_GOtable.pl

- reformat large GO annotation table (from Sam) 
- pull only GeneID & Annotation_GO_ID columns
- blank -> unknown
- csv -> .tab



