---
title: "Global Gene Expression"
author: "Megan Delatte-Guidry"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries
```{r, message=FALSE, warning=FALSE}
library(DESeq2)
library(data.table)
library(dplyr)
library(tidyr)
library(reshape2)
library(apeglm)
library(ggplot2)
library(vsn)
library(pheatmap)
library(RColorBrewer)
library(genefilter)
library(rsconnect)
library(gplots)
library(ashr)
library(limma)
```

## Load in data
```{r}
# load in metadata
sample_info <- read.csv("../de_analysis/juve_cado_metadata_w_batches.csv", header=TRUE, sep=",", row.names=1)
head(sample_info)
tail(sample_info)
```

```{r}
# load in gene count matrix and set the row names header as "gene_id"
#gene_counts <- as.data.frame(read.csv("gene_count_matrix.csv", row.names="gene_id"))
gene_counts <- read.csv("../de_analysis/gene_count_matrix.csv", row.names = 1, check.names = FALSE)
head(rownames(gene_counts))
head(gene_counts[,1:6])
```

## Check sample names
Manual: "It is absolutely critical that the columns of the count matrix and the rows of the column data (information about samples) are in the same order. DESeq2 will not make guesses as to which column of the count matrix belongs to which row of the column data, these must be provided to DESeq2 already in consistent order."  

The sample names in `sample_info` and `gene_counts` have to match up and be in the exact same order!
```{r}
# remove the "_F" in the column names of `gene_counts`
colnames(gene_counts) <- sub("_F", "", colnames(gene_counts))
head(gene_counts)

#check that the sample names are the same in both places
all(rownames(sample_info) %in% colnames(gene_counts))
#check the order of the samples
all(rownames(sample_info) == colnames(gene_counts))
```

## Subsetting data into PHASE I and PHASE II
```{r}
# subset metadata ('sample_info') first
sample_info <- tibble::rownames_to_column(sample_info, var = "sample.name")

p1_sample_info <- subset(sample_info, phase=="phase1")
p2_sample_info <- subset(sample_info, phase=="phase2")
dim(p1_sample_info)
dim(p2_sample_info)

#split count matrix based on names in each metadata variable
p1_gene_counts <- gene_counts[, p1_sample_info$sample.name]
dim(p1_gene_counts)
p2_gene_counts <- gene_counts[, p2_sample_info$sample.name]
dim(p2_gene_counts)
```


## Pre-filtering with genefilter()
While pre-filtering of low count genes isn't necessary, it can be useful for 2 reasons:    
1. reduce size of the `dds` data object (thus increase speed of count modeling).  
2. improve visualizations by making them a little cleaner


```{r}
###filtering values for PoverA
#set filter values for PoverA (P percent of the samples have counts over A)
# here P = 0.055 and A = 5 (5.5% of samples would be 4 out of 72 samples)
filt <- filterfun(pOverA(0.055,5))
#create filter for the counts data
tfil <- genefilter(gene_counts, filt)
#identify transcripts to keep by count filter
keep <- gene_counts[tfil,]
#identify gene names in transcript list to keep
gn.keep <- rownames(keep)

#data filtered in PoverA, P percent of the samples have counts over A
filt_gene_counts <- as.data.frame(gene_counts[which(rownames(gene_counts) %in% gn.keep),])
dim(gene_counts)
dim(filt_gene_counts)
#write.csv(filt_gene_counts, file="filtered_gene_count_matrix.csv")
head(filt_gene_counts,10)
```

```{r}
###filtering values for PoverA
#set filter values for PoverA (P percent of the samples have counts over A)

## PHASE I
# here P = 0.167 and A = 10 (16.7% of samples would be 4 out of 24 samples)
p1_filt <- filterfun(pOverA(0.167,10))
#create filter for the counts data
p1_tfil <- genefilter(p1_gene_counts, p1_filt)
#identify transcripts to keep by count filter
p1_keep <- p1_gene_counts[p1_tfil,]
#identify gene names in transcript list to keep
p1_gn.keep <- rownames(p1_keep)
#subset filtered genes 
p1_filt_gene_counts <- as.data.frame(p1_gene_counts[which(rownames(p1_gene_counts) %in% p1_gn.keep),])
dim(p1_gene_counts)
dim(p1_filt_gene_counts)
#write.csv(p1_filt_gene_counts, file="data/p1_filtered_gene_count_matrix.csv")
head(p1_filt_gene_counts,10)

## PHASE II
# here P = 0.083 and A = 10 (8.3% of samples would be 4 out of 48 samples)
p2_filt <- filterfun(pOverA(0.083,10))
#create filter for the counts data
p2_tfil <- genefilter(p2_gene_counts, p2_filt)
#identify transcripts to keep by count filter
p2_keep <- p2_gene_counts[p2_tfil,]
#identify gene names in transcript list to keep
p2_gn.keep <- rownames(p2_keep)
#subset filtered genes 
p2_filt_gene_counts <- as.data.frame(p2_gene_counts[which(rownames(p2_gene_counts) %in% p2_gn.keep),])
dim(p2_gene_counts)
dim(p2_filt_gene_counts)
#write.csv(p2_filt_gene_counts, file="data/p2_filtered_gene_count_matrix.csv")
head(p2_filt_gene_counts,10)
```

## Global transcriptome - all samples - both phases 

Build DeseqDataSet 
```{r}
#set batch numbers as factors 
sample_info$batch <- factor(sample_info$batch, levels = c("1", "2", "3", "4", "5"))
#sample_info$batch <- factor(sample_info$batch, levels = c("1", "2"))

# actually build the DESeq dataset
global_dds <- DESeqDataSetFromMatrix(countData = filt_gene_counts,
                             colData = sample_info, 
                             design = ~1) # ~1 means there is no design

# apply variance transformation
global_vsdds <- vst(global_dds, blind = TRUE)

#limma removeBatchEffect()
mat1 <- assay(global_vsdds)
mm1 <- model.matrix(~treatment, colData(global_vsdds))
mat1 <- limma::removeBatchEffect(mat1, batch=global_vsdds$batch, design=mm1)
assay(global_vsdds) <- mat1

# plot
plotPCA(global_vsdds, intgroup = "batch")
```
limma corrected vs non-limma corrected fancy pca
```{r}
pcaData <- plotPCA(global_vsdds,
                   intgroup = c("batch"),
                   returnData = TRUE)

percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData,
       aes(x = PC1, y = PC2,
           color = batch)) +    
  geom_point(aes(color=batch), size = 4, alpha=.75) +
  coord_fixed() +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme(
    #legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.title=element_text(face="bold", size=20, color="black"),
    axis.text=element_text(size=20, color="black"), 
    axis.text.x = element_text(size = 20),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    panel.background = element_rect(fill = alpha("white", 0.1))
    )

#ggsave("pca_plots/____", last_plot())
```

Let's get a better look with a fancier biplot
```{r}
# get PCA data from DESeq2
pcaData <- plotPCA(global_vsdds,
                   intgroup = c("treatment", "phase"),
                   returnData = TRUE)

percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData,
       aes(x = PC1, y = PC2,
           fill = treatment,    # fill by treatment
           shape = phase)) +     # shape by phase (your stress variable)
  geom_point(aes(fill=treatment), size = 4, alpha=.75, colour= "black") +
  scale_fill_manual(values = c(
    control  = "#A29BAD",
    stress   = "#D29F13",
    naive    = "#6BA84C",
    recovery = "#65A6E2"
  )) +
  scale_shape_manual(values = c(
    phase1 = 21,
    phase2 = 23
  )) +
  guides(
    fill = guide_legend(
      override.aes = list(
        shape  = 21,        # ensure fillable shape in legend
        colour = "black"    # black outline in legend keys
      )
    )
  ) +
  coord_fixed() +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme(
    #legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.title=element_text(face="bold", size=20, color="black"),
    axis.text=element_text(size=20, color="black"), 
    axis.text.x = element_text(size = 20),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    panel.background = element_rect(fill = alpha("white", 0.1))
    )

#ggsave("pca_plots/global_pca_5batch_bytreatmentphase.png", last_plot())
```


## Phase I global expression
```{r}

p1_sample_info$batch <- factor(p1_sample_info$batch, levels = c("1", "2"))

# actually build the DESeq dataset
global_dds_1 <- DESeqDataSetFromMatrix(countData = p1_filt_gene_counts,
                             colData = p1_sample_info, 
                             design = ~1) # ~1 means there is no design

# apply variance transformation
global_vsdds_1 <- vst(global_dds_1, blind = TRUE)

#limma removeBatchEffect()
mat1 <- assay(global_vsdds_1)
mm1 <- model.matrix(~treatment, colData(global_vsdds_1))
mat1 <- limma::removeBatchEffect(mat1, batch=global_vsdds_1$batch, design=mm1)
assay(global_vsdds_1) <- mat1

# plot
plotPCA(global_vsdds_1, intgroup = "treatment")
```

Let's get a better look with a fancier biplot
```{r}
# get PCA data from DESeq2
pcaData <- plotPCA(global_vsdds_1,
                   intgroup = c("treatment"),
                   returnData = TRUE)

percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData,
       aes(x = PC1, y = PC2,
           fill = treatment))+   # fill by treatment
  geom_point(aes(fill=treatment), size = 4, alpha=.75, shape=21) +
  scale_fill_manual(values = c(
    control  = "#A29BAD",
    stress   = "#D29F13"
  )) +
  coord_fixed() +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme(
    #legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.title=element_text(face="bold", size=20, color="black"),
    axis.text=element_text(size=20, color="black"), 
    axis.text.x = element_text(size = 20),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    panel.background = element_rect(fill = alpha("white", 0.1))
    )

ggsave("global_pca/p1_global_pca.png", last_plot())
```



## Phase II global expression
```{r}
#set batch numbers as factors 
p2_sample_info$batch <- factor(p2_sample_info$batch, levels = c("2", "3", "4", "5"))
#p2_sample_info$batch <- factor(p2_sample_info$batch, levels = c("1", "2"))

# build the DESeq dataset
global_dds_2 <- DESeqDataSetFromMatrix(countData = p2_filt_gene_counts,
                             colData = p2_sample_info, 
                             design = ~1) # ~1 means there is no design

# apply variance transformation
global_vsdds_2 <- vst(global_dds_2, blind = TRUE)

#limma removeBatchEffect()
mat2 <- assay(global_vsdds_2)
mm2 <- model.matrix(~treatment, colData(global_vsdds_2))
mat2 <- limma::removeBatchEffect(mat2, batch=global_vsdds_2$batch, design=mm2)
assay(global_vsdds_2) <- mat2

# plot
plotPCA(global_vsdds_2, intgroup = "treatment")
```

Let's get a better look with a fancier biplot
```{r}
# get PCA data from DESeq2
pcaData <- plotPCA(global_vsdds_2,
                   intgroup = c("treatment"),
                   returnData = TRUE)

percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData,
       aes(x = PC1, y = PC2,
           fill = treatment)) +     # 
  geom_point(aes(fill=treatment), size = 4, alpha=.75, shape = 21) +
  scale_fill_manual(values = c(
    control  = "#A29BAD",
    stress   = "#D29F13",
    naive    = "#6BA84C",
    recovery = "#65A6E2"
  )) +
  coord_fixed() +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme(
    #legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.title=element_text(face="bold", size=20, color="black"),
    axis.text=element_text(size=20, color="black"), 
    axis.text.x = element_text(size = 20),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    panel.background = element_rect(fill = alpha("white", 0.1))
    )

ggsave("global_pca/p2_global_pca.png", last_plot())
```
