---
title: "TopGO with WGCNA modules"
author: "Megan Delatte-Guidry"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Gene Ontology Enrichment Analysis on WGCNA modules

Notes****

```{r}
library(topGO)
```

### Steps

1. Data preparation: List of genes identifiers, gene scores, list of differentially expressed genes or a criteria for selecting genes based on their scores, as well as gene-to-GO annotations are all collected and stored in a single R object.   
(referenced [Amy Zyck's TopGO script](https://github.com/MarineEvoEcoLab/ROD_CADO/blob/main/Outlier_Exploration/Outlier_GO.Rmd) for the gene2go table)

2. Running the enrichment tests: Using the object created in the first step the user can perform enrichment analysis using any feasible mixture of statistical tests and methods that deal with the GO topology.

3. Analysis of the results: The results obtained in the second step are analysed using summary functions and visualisation tools.
_________________________________________________________________________________

## Step 1: Data preparation 

### gene2go table from annotation information

gaf from NCBI with the GO terms for genes in the eastern oyster genome. I'm going to take this file and extract the Gene IDs and associated GO terms.

In terminal:
```{bash}
wget https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/002/022/765/GCF_002022765.2_C_virginica-3.0/GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf.gz

gunzip GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf.gz

mv GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf gene_ontology.gaf
```

Now I can make a new .tab file with just the gene ID and GO IDs (3rd and 5th columns) renamed "gene_id" and "GO_Annotation". The gene_name column will have one unique observation per row and all associated GO terms will be saved in the same row, with values separated by ;

```{bash}
awk 'NR>9 {print $3 "\t" $5}' gene_ontology.gaf | 
sed '1i gene_id\tGO_Annotation' | 
sort | awk '
NR==1 {print; next}
$1!=prev {
if (NR>2) print prev "\t" goids;
prev=$1;
goids=$2;
next
}
{
goids=goids "; " $2
}
END {print prev "\t" goids}
' > gene2go_full.tab
```

Gene IDs for the different modules are in the following files:
phase1_outputs/phase1_module_<color>_genes.csv
phase2_outputs/phase2_module_<color>_genes.csv

I'm making another file with the gene IDs and GO IDs for the candidate loci:
```{bash}
## FORMAT THE GENE IDS & RM MSTRG??

grep -F -f phase1_outputs/phase1_module_blue_genes.csv gene2go_full.tab > p1_blue_genes_GO.tab

# Add header
awk 'BEGIN {print "gene_id\tGO_Annotation"} {print}' p1_blue_genes_GO.tab > p1_blue_genes_GO_header.tab
```

Read in gene2go information 

```{r}
all_gene2go<-read.delim("gene2go_full.tab", sep="\t")
p1_blue_gene2go<-read.delim("p1_blue_genes_GO_header.tab", sep="\t")
```

Make list of genes for input to TopGO

```{r}
cand_genes <- as.character(unique(out_gene2go$gene_name))
all_genes <- as.character(unique(all_gene2go$gene_name))
out_GeneList <- factor(as.integer(all_genes %in% cand_genes))
names(out_GeneList) <- all_genes
```

The following code will perform GO enrichment using the weighted Fisher's exact test  

Read in gene-to-go-mappings
```{r}
gene2go_topgo <-readMappings("gene2go_full.tab", IDsep=";", sep="\t")
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```

### Biological Processes

### Molecular Functions

### Cellular Components



