---
title: "TopGO with WGCNA modules"
author: "Megan Delatte-Guidry"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Gene Ontology Enrichment Analysis on WGCNA modules

Notes****

```{r, warning=FALSE, echo=FALSE}
library(topGO)
library(stringr)
library(dplyr)
library(tidyr)
```

### Steps

1. Data preparation: List of genes identifiers, gene scores, list of differentially expressed genes or a criteria for selecting genes based on their scores, as well as gene-to-GO annotations are all collected and stored in a single R object.   
(referenced [Amy Zyck's TopGO script](https://github.com/MarineEvoEcoLab/ROD_CADO/blob/main/Outlier_Exploration/Outlier_GO.Rmd) for the gene2go table)

2. Running the enrichment tests: Using the object created in the first step the user can perform enrichment analysis using any feasible mixture of statistical tests and methods that deal with the GO topology.

3. Analysis of the results: The results obtained in the second step are analysed using summary functions and visualization tools.
_________________________________________________________________________________

## Step 1: Data preparation 

### gene2go table from annotation information

gaf from NCBI with the GO terms for genes in the eastern oyster genome. I'm going to take this file and extract the Gene IDs and associated GO terms.

In terminal:
```{bash}
#wget https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/002/022/765/GCF_002022765.2_C_virginica-3.0/GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf.gz

#gunzip GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf.gz

#mv GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf gene_ontology.gaf
```

Now I can make a new .tab file with just the gene ID and GO IDs (3rd and 5th columns) renamed "gene_id" and "GO_Annotation". The gene_name column will have one unique observation per row and all associated GO terms will be saved in the same row, with values separated by ;

```{bash}
#awk 'NR>9 {print $3 "\t" $5}' gene_ontology.gaf | 
#sed '1i gene_id\tGO_Annotation' | 
#sort | awk '
#NR==1 {print; next}
#$1!=prev {
#if (NR>2) print prev "\t" goids;
#prev=$1;
#goids=$2;
#next
#}
#{
#goids=goids "; " $2
#}
#END {print prev "\t" goids}
#' > gene2go_full.tab
```

Gene IDs for the different modules are in the following files:
phase1_outputs/phase1_module_<color>_genes.csv
phase2_outputs/phase2_module_<color>_genes.csv

Make a file with gene IDs and GO IDs for the color

### Phase I BLUE
```{r}
# load data
p1_blue <- read.csv("phase1_outputs/phase1_module_blue_genes.csv", header = TRUE, sep = ",")
p1_blue
# load go annotation file to assign go terms to color module
go_terms <- read.table("gene2go_full.tab", header = TRUE, sep="\t")
head(go_terms)

# reformat gene_id column to be just LOC#, add in go annotation, remove 'MSTRG' rows and NA rows
p1_blue$gene_id <- str_extract(p1_blue$x, "[^|]+$")
dim(p1_blue)
p1_blue_wGO <- merge(p1_blue, go_terms, by = "gene_id", all.x = TRUE)
dim(p1_blue_wGO)
#clean up 
p1_blue_wGO <- na.omit(p1_blue_wGO)
p1_blue_wGO <- p1_blue_wGO[, c("gene_id", "GO_Annotation")]
dim(p1_blue_wGO)

#write.table(p1_blue_wGO, "p1_blue_genes_GO_header.tab", sep = "\t", row.names=FALSE)
```

### Phase II SEASHELL
```{r}
# load data
p2_seashell <- read.csv("phase2_outputs/phase2_module_seashell_genes.csv", header = TRUE, sep = ",")
p2_seashell
# load go annotation file to assign go terms to color module
go_terms <- read.table("gene2go_full.tab", header = TRUE, sep="\t")
head(go_terms)

# reformat gene_id column to be just LOC#, add in go annotation, remove 'MSTRG' rows and NA rows
p2_seashell$gene_id <- str_extract(p2_seashell$x, "[^|]+$")
dim(p2_seashell)
p2_seashell_wGO <- merge(p2_seashell, go_terms, by = "gene_id", all.x = TRUE)
dim(p2_seashell_wGO)
#clean up 
p2_seashell_wGO <- na.omit(p2_seashell_wGO)
p2_seashell_wGO <- p2_seashell_wGO[, c("gene_id", "GO_Annotation")]
dim(p2_seashell_wGO)

write.table(p2_seashell_wGO, "p2_seashell_genes_GO_header.tab", sep = "\t", row.names=FALSE)
```

### Phase II GOLD
```{r}
# load data
p2_gold <- read.csv("phase2_outputs/phase2_module_gold_genes.csv", header = TRUE, sep = ",")
p2_gold
# load go annotation file to assign go terms to color module
go_terms <- read.table("gene2go_full.tab", header = TRUE, sep="\t")
head(go_terms)

# reformat gene_id column to be just LOC#, add in go annotation, remove 'MSTRG' rows and NA rows
p2_gold$gene_id <- str_extract(p2_gold$x, "[^|]+$")
dim(p2_gold)
p2_gold_wGO <- merge(p2_gold, go_terms, by = "gene_id", all.x = TRUE)
dim(p2_gold_wGO)
#clean up 
p2_gold_wGO <- na.omit(p2_gold_wGO)
p2_gold_wGO <- p2_gold_wGO[, c("gene_id", "GO_Annotation")]
dim(p2_gold_wGO)

#write.table(p2_gold_wGO, "p2_gold_genes_GO_header.tab", sep = "\t", row.names=FALSE)
```

### Phase II PURPLE
```{r}
# load data
p2_purple <- read.csv("phase2_outputs/phase2_module_purple_genes.csv", header = TRUE, sep = ",")
p2_purple
# load go annotation file to assign go terms to color module
go_terms <- read.table("gene2go_full.tab", header = TRUE, sep="\t")
head(go_terms)

# reformat gene_id column to be just LOC#, add in go annotation, remove 'MSTRG' rows and NA rows
p2_purple$gene_id <- str_extract(p2_purple$x, "[^|]+$")
dim(p2_purple)
p2_purple_wGO <- merge(p2_purple, go_terms, by = "gene_id", all.x = TRUE)
dim(p2_purple_wGO)
#clean up 
p2_purple_wGO <- na.omit(p2_purple_wGO)
p2_purple_wGO <- p2_purple_wGO[, c("gene_id", "GO_Annotation")]
dim(p2_purple_wGO)

#write.table(p2_purple_wGO, "p2_purple_genes_GO_header.tab", sep = "\t", row.names=FALSE)
```

Read in expressed genes file - maybe I'll use this??
```{r}
p1_filt_counts <- read.csv("../de_analysis/outputs/p1_filtered_gene_count_matrix.csv")
head(p1_filt_counts)
```


Read in gene2go information 

```{r}
all_gene2go<-read.delim("gene2go_full.tab", sep="\t")
module_gene2go<-read.delim("p2_purple_genes_GO_header.tab", sep="\t")
View(module_gene2go)
```

Make list of genes for input to TopGO

```{r}
mod_genes <- as.character(unique(module_gene2go$gene_id))
all_genes <- as.character(unique(all_gene2go$gene_id))
mod_GeneList <- factor(as.integer(all_genes %in% mod_genes))
names(mod_GeneList) <- all_genes
```

The following code will perform GO enrichment using the weighted Fisher's exact test  

Read in gene-to-go-mappings
```{r}
gene2go_topgo <-readMappings("gene2go_full.tab", IDsep=";", sep="\t")
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```

### Biological Processes

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_OUT_BP_pi <-new("topGOdata", ontology="BP", gene2GO=gene2go_topgo, allGenes=mod_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_OUT_BP_FE_pi <- runTest(GO_OUT_BP_pi, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_OUT_BP_En_pi <- GenTable(GO_OUT_BP_pi, Fisher = GO_OUT_BP_FE_pi, orderBy = "Fisher",  topNodes = 50, numChar = 100)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_OUT_BP_En_pi$Fisher<-as.numeric(GO_OUT_BP_En_pi$Fisher)
GO_OUT_BP_En_sig_pi<-GO_OUT_BP_En_pi[GO_OUT_BP_En_pi$Fisher<0.05,]
```

Merge `GO_OUT_BP_En_sig_pi` with GO and gene info. 
```{r}
# Separate GO terms 
module_gene2go <- module_gene2go %>%
  separate_rows(GO_Annotation, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
module_gene2go$GO_Annotation <- trimws(module_gene2go$GO_Annotation)
GO_OUT_BP_En_sig_pi$GO.ID <- trimws(GO_OUT_BP_En_sig_pi$GO.ID)

# Join the datasets based on GO term
GO_OUT_BP_En_sig_gene_pi <- module_gene2go %>%
  left_join(GO_OUT_BP_En_sig_pi, by = c("GO_Annotation" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_OUT_BP_En_sig_gene_pi$ontology <- "Biological Processes"
```

### Cellular Components
Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_OUT_CC_pi <-new("topGOdata", ontology="CC", gene2GO=gene2go_topgo, allGenes=mod_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_OUT_CC_FE_pi <- runTest(GO_OUT_CC_pi, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_OUT_CC_En_pi <- GenTable(GO_OUT_CC_pi, Fisher = GO_OUT_CC_FE_pi, orderBy = "Fisher",  topNodes = 50, numChar = 100)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_OUT_CC_En_pi$Fisher<-as.numeric(GO_OUT_CC_En_pi$Fisher)
GO_OUT_CC_En_sig_pi<-GO_OUT_CC_En_pi[GO_OUT_CC_En_pi$Fisher<0.05,]
```

Merge `GO_OUT_CC_En_sig_pi` with GO and gene info. 
```{r}
# Separate GO terms 
module_gene2go <- module_gene2go %>%
  separate_rows(GO_Annotation, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
module_gene2go$GO_Annotation <- trimws(module_gene2go$GO_Annotation)
GO_OUT_CC_En_sig_pi$GO.ID <- trimws(GO_OUT_CC_En_sig_pi$GO.ID)

# Join the datasets based on GO term
GO_OUT_CC_En_sig_gene_pi <- module_gene2go %>%
  left_join(GO_OUT_CC_En_sig_pi, by = c("GO_Annotation" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_OUT_CC_En_sig_gene_pi$ontology <- "Cellular Components"
```


### Molecular Functions

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_OUT_MF_pi <-new("topGOdata", ontology="MF", gene2GO=gene2go_topgo, allGenes=mod_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
```

Run GO enrichment test 
```{r}
GO_OUT_MF_FE_pi <- runTest(GO_OUT_MF_pi, algorithm="weight01", statistic="fisher")
```

Generate results table 
```{r}
GO_OUT_MF_En_pi <- GenTable(GO_OUT_MF_pi, Fisher = GO_OUT_MF_FE_pi, orderBy = "Fisher",  topNodes = 50, numChar = 100)
```
Only taking the top 100 GO terms 

Filter by significant results
```{r}
GO_OUT_MF_En_pi$Fisher<-as.numeric(GO_OUT_MF_En_pi$Fisher)
GO_OUT_MF_En_sig_pi<-GO_OUT_MF_En_pi[GO_OUT_MF_En_pi$Fisher<0.05,]
```

Merge `GO_OUT_MF_En_sig_pi` with GO and gene info. 
```{r}
# Separate GO terms 
module_gene2go <- module_gene2go %>%
  separate_rows(GO_Annotation, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
module_gene2go$GO_Annotation <- trimws(module_gene2go$GO_Annotation)
GO_OUT_MF_En_sig_pi$GO.ID <- trimws(GO_OUT_MF_En_sig_pi$GO.ID)

# Join the datasets based on GO term
GO_OUT_MF_En_sig_gene_pi <- module_gene2go %>%
  left_join(GO_OUT_MF_En_sig_pi, by = c("GO_Annotation" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_OUT_MF_En_sig_gene_pi$ontology <- "Molecular Functions"
```

### Join ontologies

Bind `GO_OUT_BP_En_sig_gene_pi`, `GO_OUT_CC_En_sig_gene_pi`, and `GO_OUT_MF_En_sig_gene_pi` to make a df that has significantly enriched GO terms for all ontologies 
```{r}
GO_OUT_En_sig_gene_pi <- rbind(GO_OUT_BP_En_sig_gene_pi, GO_OUT_CC_En_sig_gene_pi, GO_OUT_MF_En_sig_gene_pi)
```

Merge `pi_bed_filt` with `GO_OUT_En_sig_gene_pi` and calculate proportion of significant genes
```{r}
merged_out_genes <- GO_OUT_En_sig_gene_pi %>%
  mutate(prop.sig.genes = Significant/Annotated)

# saving unique instances of gene/GO pairs
merged_out_genes <- unique(merged_out_genes)
```

```{r}
# Write to csv 
#write.csv(merged_out_genes, "module_GO_enrichment/p2_purple_genes_GO_enrichment.csv")
```



Plot - need to work out 
```{r}
GO_plot_pi <- ggplot(merged_out_genes, aes(x = Term, y = -log(Fisher), size = prop.sig.genes, fill = -log(Fisher))) +
  expand_limits(y = 1.5) +
  ylim(1, 7.25) +
  #Add horizontal lines with a single aesthetic value
  geom_hline(yintercept = -log10(0.01), linetype = "longdash", colour = "black", linewidth = .6) +
  geom_hline(yintercept = -log10(0.001), linetype = "solid", colour = "black", linewidth = .6) +
  geom_point(shape = 21) + 
  scale_size(range = c(2, 12)) + 
  scale_fill_continuous(low = "#1AD3D1FF", high = "#4686FBFF") +
  xlab('') + 
  ylab('Enrichment score') +
  #labs(caption = 'Cut-off lines at p=0.01 and p=0.001') +
  theme_bw() +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); GO_plot_pi

ggsave(
  "GO_plot_pi.png",
  plot = GO_plot_pi,
  width = 8,
  height = 10,
  dpi = 300
)

```

Lastly, I'm just going to save the figure as a .png

```{r}
# Open PNG device with high resolution and transparent background
png(filename = "GO_plot_pi.png",
    width = 2000,   # width in pixels (adjust for desired size)
    height = 2000,  # height in pixels
    res = 200,      # resolution in dpi for high quality
    bg = "transparent")  # transparent background

# Print the plot object
print(GO_plot_pi)

# Close the device to finalize the file
dev.off()
```



