---
title: "TopGO with WGCNA modules"
author: "Megan Delatte-Guidry"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Gene Ontology Enrichment Analysis on WGCNA modules
```{r, warning=FALSE, echo=FALSE}
library(topGO)
library(stringr)
library(dplyr)
library(tidyr)
```

### Steps

1. Data preparation: List of genes identifiers, gene scores, list of differentially expressed genes or a criteria for selecting genes based on their scores, as well as gene-to-GO annotations are all collected and stored in a single R object.   
(referenced [Amy Zyck's TopGO script](https://github.com/MarineEvoEcoLab/ROD_CADO/blob/main/Outlier_Exploration/Outlier_GO.Rmd) for the gene2go table)

2. Running the enrichment tests: Using the object created in the first step the user can perform enrichment analysis using any feasible mixture of statistical tests and methods that deal with the GO topology.

3. Analysis of the results: The results obtained in the second step are analysed using summary functions and visualization tools.
_________________________________________________________________________________

## Step 1: Data preparation 

### gene2go table from annotation information

gaf from NCBI with the GO terms for genes in the eastern oyster genome. I'm going to take this file and extract the Gene IDs and associated GO terms.

In terminal:
```{bash}
#wget https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/002/022/765/GCF_002022765.2_C_virginica-3.0/GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf.gz

#gunzip GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf.gz

#mv GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf gene_ontology.gaf
```

Now I can make a new .tab file with just the gene ID and GO IDs (3rd and 5th columns) renamed "gene_id" and "GO_Annotation". The gene_name column will have one unique observation per row and all associated GO terms will be saved in the same row, with values separated by ;
```{bash}
#awk 'NR>9 {print $3 "\t" $5}' gene_ontology.gaf | 
#sed '1i gene_id\tGO_Annotation' | 
#sort | awk '
#NR==1 {print; next}
#$1!=prev {
#if (NR>2) print prev "\t" goids;
#prev=$1;
#goids=$2;
#next
#}
#{
#goids=goids "; " $2
#}
#END {print prev "\t" goids}
#' > gene2go_full.tab
```

Gene IDs for the different modules are in the following files:
module_lists/module_<color>_genes.csv

## Make a file with gene IDs and GO IDs for the color

### GOLD
```{r}
# load data
gold <- read.csv("module_lists/module_gold_genes.csv", header = TRUE, sep = ",")
gold
# load go annotation file to assign go terms to color module
go_terms <- read.table("gene2go_full.tab", header = TRUE, sep="\t")
head(go_terms)

# reformat gene_id column to be just LOC#, add in go annotation, remove 'MSTRG' rows and NA rows
gold$gene_id <- str_extract(gold$x, "[^|]+$")
dim(gold)
gold_wGO <- merge(gold, go_terms, by = "gene_id", all.x = TRUE)
dim(gold_wGO)
#clean up 
gold_wGO <- na.omit(gold_wGO)
gold_wGO <- gold_wGO[, c("gene_id", "GO_Annotation")]
dim(gold_wGO)

write.table(gold_wGO, "module_goterms/gold_genes_GO_header.tab", sep = "\t", row.names=FALSE)
```

### LEMON
```{r}
# load data
lemon <- read.csv("module_lists/module_lemon_genes.csv", header = TRUE, sep = ",")
dim(lemon)
# load go annotation file to assign go terms to color module
go_terms <- read.table("gene2go_full.tab", header = TRUE, sep="\t")
dim(go_terms); colnames(go_terms)

# reformat gene_id column to be just LOC#, add in go annotation, remove 'MSTRG' rows and NA rows
lemon$gene_id <- str_extract(lemon$x, "[^|]+$")
dim(lemon)
lemon_wGO <- merge(lemon, go_terms, by = "gene_id", all.x = TRUE)
dim(lemon_wGO)
#clean up 
lemon_wGO <- na.omit(lemon_wGO)
lemon_wGO <- lemon_wGO[, c("gene_id", "GO_Annotation")]
dim(lemon_wGO)

write.table(lemon_wGO, "module_goterms/lemon_genes_GO_header.tab", sep = "\t", row.names=FALSE)
```

### SAGE
```{r}
# load data
sage <- read.csv("module_lists/module_sage_genes.csv", header = TRUE, sep = ",")
dim(sage)
# load go annotation file to assign go terms to color module
go_terms <- read.table("gene2go_full.tab", header = TRUE, sep="\t")
dim(go_terms); colnames(go_terms)

# reformat gene_id column to be just LOC#, add in go annotation, remove 'MSTRG' rows and NA rows
sage$gene_id <- str_extract(sage$x, "[^|]+$")
dim(sage)
sage_wGO <- merge(sage, go_terms, by = "gene_id", all.x = TRUE)
dim(sage_wGO)
#clean up 
sage_wGO <- na.omit(sage_wGO)
sage_wGO <- sage_wGO[, c("gene_id", "GO_Annotation")]
dim(sage_wGO)

write.table(sage_wGO, "module_goterms/sage_genes_GO_header.tab", sep = "\t", row.names=FALSE)
```

### MINT
```{r}
# load data
mint <- read.csv("module_lists/module_mint_genes.csv", header = TRUE, sep = ",")
dim(mint)
# load go annotation file to assign go terms to color module
go_terms <- read.table("gene2go_full.tab", header = TRUE, sep="\t")
dim(go_terms); colnames(go_terms)

# reformat gene_id column to be just LOC#, add in go annotation, remove 'MSTRG' rows and NA rows
mint$gene_id <- str_extract(mint$x, "[^|]+$")
dim(mint)
mint_wGO <- merge(mint, go_terms, by = "gene_id", all.x = TRUE)
dim(mint_wGO)
#clean up 
mint_wGO <- na.omit(mint_wGO)
mint_wGO <- mint_wGO[, c("gene_id", "GO_Annotation")]
dim(mint_wGO)

write.table(mint_wGO, "module_goterms/mint_genes_GO_header.tab", sep = "\t", row.names=FALSE)
```

### FOREST
```{r}
# load data
forest <- read.csv("module_lists/module_forest_genes.csv", header = TRUE, sep = ",")
dim(forest)
# load go annotation file to assign go terms to color module
go_terms <- read.table("gene2go_full.tab", header = TRUE, sep="\t")
dim(go_terms); colnames(go_terms)

# reformat gene_id column to be just LOC#, add in go annotation, remove 'MSTRG' rows and NA rows
forest$gene_id <- str_extract(forest$x, "[^|]+$")
dim(forest)
forest_wGO <- merge(forest, go_terms, by = "gene_id", all.x = TRUE)
dim(forest_wGO)
#clean up 
forest_wGO <- na.omit(forest_wGO)
forest_wGO <- forest_wGO[, c("gene_id", "GO_Annotation")]
dim(forest_wGO)

write.table(forest_wGO, "module_goterms/forest_genes_GO_header.tab", sep = "\t", row.names=FALSE)
```

### LILAC
```{r}
# load data
lilac <- read.csv("module_lists/module_lilac_genes.csv", header = TRUE, sep = ",")
dim(lilac)
# load go annotation file to assign go terms to color module
go_terms <- read.table("gene2go_full.tab", header = TRUE, sep="\t")
dim(go_terms); colnames(go_terms)

# reformat gene_id column to be just LOC#, add in go annotation, remove 'MSTRG' rows and NA rows
lilac$gene_id <- str_extract(lilac$x, "[^|]+$")
dim(lilac)
lilac_wGO <- merge(lilac, go_terms, by = "gene_id", all.x = TRUE)
dim(lilac_wGO)
#clean up 
lilac_wGO <- na.omit(lilac_wGO)
lilac_wGO <- lilac_wGO[, c("gene_id", "GO_Annotation")]
dim(lilac_wGO)

write.table(lilac_wGO, "module_goterms/lilac_genes_GO_header.tab", sep = "\t", row.names=FALSE)
```

### IRIS
```{r}
# load data
iris <- read.csv("module_lists/module_iris_genes.csv", header = TRUE, sep = ",")
dim(iris)
# load go annotation file to assign go terms to color module
go_terms <- read.table("gene2go_full.tab", header = TRUE, sep="\t")
dim(go_terms); colnames(go_terms)

# reformat gene_id column to be just LOC#, add in go annotation, remove 'MSTRG' rows and NA rows
iris$gene_id <- str_extract(iris$x, "[^|]+$")
dim(iris)
iris_wGO <- merge(iris, go_terms, by = "gene_id", all.x = TRUE)
dim(iris_wGO)
#clean up 
iris_wGO <- na.omit(iris_wGO)
iris_wGO <- iris_wGO[, c("gene_id", "GO_Annotation")]
dim(iris_wGO)

write.table(iris_wGO, "module_goterms/iris_genes_GO_header.tab", sep = "\t", row.names=FALSE)
```

### PLUM
```{r}
# load data
plum <- read.csv("module_lists/module_plum_genes.csv", header = TRUE, sep = ",")
dim(plum)
# load go annotation file to assign go terms to color module
go_terms <- read.table("gene2go_full.tab", header = TRUE, sep="\t")
dim(go_terms); colnames(go_terms)

# reformat gene_id column to be just LOC#, add in go annotation, remove 'MSTRG' rows and NA rows
plum$gene_id <- str_extract(plum$x, "[^|]+$")
dim(plum)
plum_wGO <- merge(plum, go_terms, by = "gene_id", all.x = TRUE)
dim(plum_wGO)
#clean up 
plum_wGO <- na.omit(plum_wGO)
plum_wGO <- plum_wGO[, c("gene_id", "GO_Annotation")]
dim(plum_wGO)

write.table(plum_wGO, "module_goterms/plum_genes_GO_header.tab", sep = "\t", row.names=FALSE)
```



## TOPGO ANALYSIS 
Read in gene2go information 
```{r}
# SET MODULE COLOR 
color <- "plum"
module_gene2go<-read.delim(paste0("module_goterms/", color,"_genes_GO_header.tab"), sep="\t")
View(module_gene2go)

# ALL ANNOTATED GENE ONTOLOGY INFO
all_gene2go<-read.delim("gene2go_full.tab", sep="\t")
```

Make list of genes for input to TopGO

```{r}
mod_genes <- as.character(unique(module_gene2go$gene_id))
all_genes <- as.character(unique(all_gene2go$gene_id))
mod_GeneList <- factor(as.integer(all_genes %in% mod_genes))
names(mod_GeneList) <- all_genes
```

The following code will perform GO enrichment using the weighted Fisher's exact test  

Read in gene-to-go-mappings
```{r}
gene2go_topgo <-readMappings("gene2go_full.tab", IDsep=";", sep="\t")
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```

### Biological Processes
```{r}
#Create `topGOdata` object, which is required for topGO analysis
GO_OUT_BP_pi <-new("topGOdata", ontology="BP", gene2GO=gene2go_topgo, allGenes=mod_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)

#Run GO enrichment test 
GO_OUT_BP_FE_pi <- runTest(GO_OUT_BP_pi, algorithm="weight01", statistic="fisher")

#Generate results table 
GO_OUT_BP_En_pi <- GenTable(GO_OUT_BP_pi, Fisher = GO_OUT_BP_FE_pi, orderBy = "Fisher",  topNodes = 50, numChar = 100)

#Only taking the top 100 GO terms 

#Filter by significant results
GO_OUT_BP_En_pi$Fisher<-as.numeric(GO_OUT_BP_En_pi$Fisher)
GO_OUT_BP_En_sig_pi<-GO_OUT_BP_En_pi[GO_OUT_BP_En_pi$Fisher<0.05,]


#Merge `GO_OUT_BP_En_sig_pi` with GO and gene info. 
# Separate GO terms 
module_gene2go <- module_gene2go %>%
  separate_rows(GO_Annotation, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
module_gene2go$GO_Annotation <- trimws(module_gene2go$GO_Annotation)
GO_OUT_BP_En_sig_pi$GO.ID <- trimws(GO_OUT_BP_En_sig_pi$GO.ID)

# Join the datasets based on GO term
GO_OUT_BP_En_sig_gene_pi <- module_gene2go %>%
  left_join(GO_OUT_BP_En_sig_pi, by = c("GO_Annotation" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_OUT_BP_En_sig_gene_pi$ontology <- "Biological Processes"
```

### Cellular Components
```{r}
#Create `topGOdata` object, which is required for topGO analysis
GO_OUT_CC_pi <-new("topGOdata", ontology="CC", gene2GO=gene2go_topgo, allGenes=mod_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)

#Run GO enrichment test 
GO_OUT_CC_FE_pi <- runTest(GO_OUT_CC_pi, algorithm="weight01", statistic="fisher")

#Generate results table 
GO_OUT_CC_En_pi <- GenTable(GO_OUT_CC_pi, Fisher = GO_OUT_CC_FE_pi, orderBy = "Fisher",  topNodes = 50, numChar = 100)
#Only taking the top 100 GO terms 

#Filter by significant results
GO_OUT_CC_En_pi$Fisher<-as.numeric(GO_OUT_CC_En_pi$Fisher)
GO_OUT_CC_En_sig_pi<-GO_OUT_CC_En_pi[GO_OUT_CC_En_pi$Fisher<0.05,]


#Merge `GO_OUT_CC_En_sig_pi` with GO and gene info. 
# Separate GO terms 
module_gene2go <- module_gene2go %>%
  separate_rows(GO_Annotation, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
module_gene2go$GO_Annotation <- trimws(module_gene2go$GO_Annotation)
GO_OUT_CC_En_sig_pi$GO.ID <- trimws(GO_OUT_CC_En_sig_pi$GO.ID)

# Join the datasets based on GO term
GO_OUT_CC_En_sig_gene_pi <- module_gene2go %>%
  left_join(GO_OUT_CC_En_sig_pi, by = c("GO_Annotation" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_OUT_CC_En_sig_gene_pi$ontology <- "Cellular Components"
```


### Molecular Functions

Create `topGOdata` object, which is required for topGO analysis
```{r}
GO_OUT_MF_pi <-new("topGOdata", ontology="MF", gene2GO=gene2go_topgo, allGenes=mod_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)

#Run GO enrichment test 
GO_OUT_MF_FE_pi <- runTest(GO_OUT_MF_pi, algorithm="weight01", statistic="fisher")

#Generate results table 
GO_OUT_MF_En_pi <- GenTable(GO_OUT_MF_pi, Fisher = GO_OUT_MF_FE_pi, orderBy = "Fisher",  topNodes = 50, numChar = 100)
#Only taking the top 100 GO terms 

#Filter by significant results
GO_OUT_MF_En_pi$Fisher<-as.numeric(GO_OUT_MF_En_pi$Fisher)
GO_OUT_MF_En_sig_pi<-GO_OUT_MF_En_pi[GO_OUT_MF_En_pi$Fisher<0.05,]

#Merge `GO_OUT_MF_En_sig_pi` with GO and gene info. 
# Separate GO terms 
module_gene2go <- module_gene2go %>%
  separate_rows(GO_Annotation, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
module_gene2go$GO_Annotation <- trimws(module_gene2go$GO_Annotation)
GO_OUT_MF_En_sig_pi$GO.ID <- trimws(GO_OUT_MF_En_sig_pi$GO.ID)

# Join the datasets based on GO term
GO_OUT_MF_En_sig_gene_pi <- module_gene2go %>%
  left_join(GO_OUT_MF_En_sig_pi, by = c("GO_Annotation" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_OUT_MF_En_sig_gene_pi$ontology <- "Molecular Functions"
```

### Join ontologies

Bind `GO_OUT_BP_En_sig_gene_pi`, `GO_OUT_CC_En_sig_gene_pi`, and `GO_OUT_MF_En_sig_gene_pi` to make a df that has significantly enriched GO terms for all ontologies 
```{r}
GO_OUT_En_sig_gene_pi <- rbind(GO_OUT_BP_En_sig_gene_pi, GO_OUT_CC_En_sig_gene_pi, GO_OUT_MF_En_sig_gene_pi)
```

Merge `pi_bed_filt` with `GO_OUT_En_sig_gene_pi` and calculate proportion of significant genes
```{r}
merged_out_genes <- GO_OUT_En_sig_gene_pi %>%
  mutate(prop.sig.genes = Significant/Annotated)

# saving unique instances of gene/GO pairs
merged_out_genes <- unique(merged_out_genes)
```

Write out enriched GO terms for the given module color
```{r}
# Write to csv 
write.csv(merged_out_genes, paste0("module_go_enrichment/",color,"_genes_GO_enrichment.csv"))
```


## Visualization Lagniappe

Set color and read in GO enrichment data for plotting below
```{r}
color <- "plum"
merged_out_genes <- read.csv(paste0("module_go_enrichment/",color,"_genes_GO_enrichment.csv"))
```

Bubble plot - update colors!
```{r}
GO_plot_pi <- ggplot(merged_out_genes, aes(x = Term, y = -log(Fisher), size = Annotated, fill = -log(Fisher))) +
  expand_limits(y = 1.5) +
  #ylim(1, 40) +
  #Add horizontal lines with a single aesthetic value
  geom_hline(yintercept = -log(0.01), linetype = "longdash", colour = "black", linewidth = .6) +
  geom_hline(yintercept = -log(0.001), linetype = "solid", colour = "black", linewidth = .6) +
  geom_point(shape = 21) + 
  scale_size(range = c(2, 8)) + 
  #scale_fill_continuous(low = "goldenrod1", high = "goldenrod4") + #gold
  #scale_fill_continuous(low = "khaki", high = "khaki4") + #lemon
  #scale_fill_continuous(low = "honeydew", high = "honeydew4") + #sage
  #scale_fill_continuous(low = "palegreen", high = "palegreen4") + #mint
  #scale_fill_continuous(low = "seagreen", high = "#1E5938") + #forest
  #scale_fill_continuous(low = "thistle", high = "thistle4") + #lilac
  #scale_fill_continuous(low = "mediumpurple", high = "mediumpurple4") + #iris
  #scale_fill_continuous(low = "plum4", high = "#5C3A5C") + #plum
  xlab('') + 
  ylab('Enrichment score') +
  labs(caption = 'Cut-off lines at p=0.01 (dashed) and p=0.001 (solid)', 
       title = paste0("Module ", color, " enriched GO terms")) +
  theme_bw() +
  facet_grid(vars(ontology), scales = "free", space = "free_y") +
  coord_flip(); GO_plot_pi

ggsave(
  paste0("bubble_plots/",color,"_GO_bubbleplot.png"),
  plot = GO_plot_pi,
  width = 8,
  height = 6,
  dpi = 300
)
```


Looking for overlaps between enriched GO terms 
```{r}

```



