---
title: "WGCNA Juve CADO"
author: "Megan Delatte-Guidry"
date: "`r Sys.Date()`"
output: github_document
---

# Weighed Gene Co-Network Analysis (WGCNA) for Juvenile CADO RNAseq project

IMPORTANT TO CONSIDER FOR RE-ANALYSIS:
- generate filtered dataset with both phases included
- reorganize beginning of script (steps 1 & 2)
- set model to accomodate two phase design 


## Steps: 
1. [Load packages and data](#1-load-packages-and-data)
2. [Pre-processing data]()
3. [Determine outliers with cluster tree]()
4. [Cut outliers and replot tree]()
5. [Update matrices for following analyses]()
6. [Prepare trait data]()
7. [Cluster samples by trait]()
8. [Set softpower threshold]()
9. [Stepwise module construction]()
10. [Prepare for module trait associations]()
11. [Module trait coorelation]()
12. [Hub genes]()
13. [Heatmaps]()

------------------------------------------------------------------------

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Load packages and data

Load packages

```{r, results="hide"}
library(WGCNA) # Jon installed version 1.63 on KITT R version 4.3.2 (2023-10-31)
library(dplyr)
library(zoo)
library(DESeq2)
#library(devtools)
library(ComplexHeatmap)
library(circlize)
library(reshape)
library(ggplot2)
#library(hrbrthemes)
```

The following setting is important, do not omit. (as recommended by WGCNA authors - view tutorial)

```{r}
options(stringsAsFactors = FALSE)
```

## 2. Pre-processing data

*Preliminary filtering of the counts with P over A*\
I used the filtered gene count matrix file I created with genefilter() in R version 3.6.0 with `de_analysis/DESeq2_analysis.Rmd`\
I removed all rows that have less than 5.5% of samples with less than 5 counts (5.5% would be 4 out of 72 samples)

While pre-filtering of low count genes isn't necessary, it can be useful for 2 reasons:\
1. reduce size of the data object (thus increase speed of modeling).\
2. improve visualizations by making them a little cleaner

[Read here for the pre processing steps using WGCNA!](https://bmcbioinformatics.biomedcentral.com/track/pdf/10.1186/1471-2105-9-559.pdf)

*Read in data*\
```{r}
# read in treatment metadata 
treatment_data <- read.csv(file="../de_analysis/juve_cado_metadata_w_batches.csv", sep=',', header=TRUE, row.names = 1)
dim(treatment_data)

# read in filtered count data 
## These data have been filtered to exclude any gene that has less than 10 counts 
count_data <- as.data.frame(read.csv("../de_analysis/wgcna_over10_filtered_gene_count_matrix.csv", row.names = 1, check.names = FALSE))
dim(count_data)

#check that the sample names are the same in both places
all(rownames(treatment_data) %in% colnames(count_data))
#check the order of the samples
all(rownames(treatment_data) == colnames(count_data))

# change row.names of treatment_data to the first column called sample.names
treatment_data <- tibble::rownames_to_column(treatment_data, var = "sample.name")
head(treatment_data)
```

> *Notes on data normalization!* The question of how to normalize this dataset came up. The creators of WGCNA insist that "properly normalized" data is required. It turns out transforming your data (ex vst or log) is a type of normalization method.

> Here, I'll use a variance stabilizing transformation on just the raw counts. "Many common statistical methods for exploratory analysis of multidimensional data, especially methods for clustering and ordination (e.g., principal-component analysis and the like), work best for (at least approximately) homoskedastic data; this means that the variance of an observable quantity (i.e., here, the expression strength of a gene) does not depend on the mean. In RNA-Seq data, however, variance grows with the mean." <https://www.biostars.org/p/109104/> <http://www.nxn.se/valent/2017/10/15/variance-stabilizing-scrna-seq-counts> The count data is in no way normally distributed "The process of counting implies that variation will propagate as the number of events increase. The effect of this is that there will be an inherent relation between mean (expected value) and variance of counts" From what I can gather the vst transformation removes the relationship between the mean and the variance.

> **We will use a variance stabilizing transformation here (vst transformation)** This transformation is recommended in the R documentation for using WGCNA on RNAseq data <https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/faq.html/>

#### Formatting data in DESeq matrix and perform vst 

Create 'dds' object (DESeqDataSet) & vst transform the 'dds' object for WGCNA (following matrix construction laid out in `de_analysis/DESeq2_analysis.Rmd`) Then we'll transform the data and select only the top most variable genes to look at in the analysis

```{r}
# assign the batch numbers as factors not integers
treatment_data$batch <- factor(treatment_data$batch, levels = c("1", "2", "3", "4", "5"))

#build deseq matrix (dm#)
dm <- DESeqDataSetFromMatrix(countData = count_data,
                             colData = treatment_data,
                             design = ~ batch + phase.treat)
dm$phase.treat <- factor(dm$phase.treat, levels = c("1_control", "1_stress", "2_control", "2_stress", "2_naive", "2_recovery"))
levels(dm$phase.treat)
dm

# transform the data  ========================================================== #
# run in order (kept name as dm_vst)
dm_vst <- vst(dm) # transform it vst (variance stabilized transformation)
dm_vst <- assay(dm_vst) # call only the transformed counts in the dds object
dim(dm_vst)
dm_vst_t <- t(dm_vst)
dim(dm_vst_t)

#############
## OPTIONAL FOR DATA REDUCTION - NOT USING IN FAVOR OF COUNT BASED FILTERING
# lets pair down to look at the top 10,000 most variable genes instead of the list of 68,435 genes
#MostVaryGenes <- head(order(rowVars(dm_vst), decreasing = TRUE), 10000)
# make that information into a matrix
#VMostVaryGenes_Mat <- dm_vst[ MostVaryGenes, ] #which in the DESeq2 style matrix are those genes?
#dim(VMostVaryGenes_Mat)
# make that matrix a dataframe
#dm_vst_5000 <- as.data.frame(VMostVaryGenes_Mat)
#dim(dm_vst_5000)
# transpose columns to rows and vice versa
#dm_mvg <- t(dm_vst_5000) 
#dim(dm_mvg)
```

## 3. Determine outliers with cluster tree

WGCNA Sample tree - omit outlier sample(s)
```{r}
# checks before we start....
dim(dm_vst_t) #  72 samples; 14,177 genes
gsg = goodSamplesGenes(dm_vst_t, verbose = 3);  # We first check for genes and samples with too many missing values:
gsg$allOK # If the statement returns TRUE, all genes have passed the cuts. 
table(gsg$goodGenes)
table(gsg$goodSamples)
```

Plot cluster tree and look for potential outliers - look at the plotted tree!!
```{r}
# call the cluster and set window dimensions to view..
sampleTree = hclust(dist(dm_vst_t), method = "average") # Next we cluster the samples (in contrast to clustering genes that will come later)  to see if there are any obvious outliers.
options(bitmapType = 'cairo')
sizeGrWindow(12,9) # The user should change the dimensions if the window is too large or too small.
par(cex = 0.6);
par(mar = c(0,4,2,0))
# output the tree as .png file..
png("outputs/sampleClusterTree_Precut.png", 1000, 800, pointsize=18)
precut_tree <- plot(sampleTree, main = "Sample clustering to detect outliers - WGCNA", sub="", xlab="", cex.lab = 1.5, 
     cex.axis = 1.5, cex.main = 2)+ abline(h = 115 , col = "red");precut_tree # add line to plot to show the cut-off outlier samples 
dev.off()
```

Plot PCA of samples w/ outliers in red
```{r}
pca <- prcomp(dm_vst_t)
pca.data <- pca$x

pca.var <- pca$sdev^2
pca.var.percent <- round(pca.var/sum(pca.var)*100, digits = 2)

pca.data <- as.data.frame(pca.data)
pca.data$SampleName <- rownames(pca.data)

pca.data$Outliers <- ifelse(pca.data$SampleName %in% c("J1_14SC2_4", "J2_14SC2_2", "J2_14SN2_1", "J2_14SN2_2", "J2_14SN3_3","J2_14SR1_2", "J2_14SR1_4", "J2_14SS1_4"), 
                           as.character(pca.data$SampleName), "Other")

png("outputs/outliers_pca.png", 800, 600)
out_pca <- ggplot(pca.data, aes(PC1, PC2, color = Outliers)) +
  #geom_point(size=2) +
  scale_color_manual(values = c("J1_14SC2_4" = "red",
                                "J2_14SC2_2"= "red",
                                "J2_14SN2_1"= "red",
                                "J2_14SN2_2"= "red",
                                "J2_14SN3_3"= "red",
                                "J2_14SR1_2"= "red",
                                "J2_14SR1_4"= "red",
                                "J2_14SS1_4"= "red",
                                "Other"= "grey10")) +
  geom_text(label = rownames(pca.data), size = 4) +
  labs(title = "PCA of samples w/ cluster tree outliers in red - WGCNA",
       x= paste0('PC1: ', pca.var.percent[1], ' %'),
       y= paste0('PC2: ', pca.var.percent[2], ' %')) +
  theme_bw() + 
  theme(legend.position = "none",
        axis.text = element_text(size = 16),
        axis.title = element_text(size = 16),
        title = element_text(face = "bold", size = 18)); out_pca
dev.off()
```

## 4. Cut outliers and replot tree

##**I'll cut outliers (1 PI control; 1 PII control; 1 PII stress; 3 PII naive; 2 PII recovery)**

Cutting the outlier(s) and re-plotting the tree. 
```{r}
# cut the tree and omit  ========================================================== #
clust = cutreeStatic(sampleTree, cutHeight = 115, minSize = 10) # Determine cluster under the line
table(clust) # 0 = cut; 1 = kept; says it will cut 8 and save 64; exactly what we want!
keepSamples = (clust==1) # 'keepsamples' boolean to call the main dataset - notice there are occurrences of FALSE - these are the samples I am cutting

# integrate keepsamples  ========================================================== #
dm_vst_t = dm_vst_t[keepSamples, ] # integreat the boolean 'keepsamples' to ommit oultilers determined in the sample tree above
nGenes = ncol(dm_vst_t) # number of genes == 14177
nSamples = nrow(dm_vst_t) # number of samples == 64  - the cut tree removed 8 sample
```

```{r}
# plot the tree with the 'keep samples'  =========================================== #
sampleTree2 = hclust(dist(dm_vst_t), method = "average") # Next we cluster the samples (in contrast to clustering genes that will come later)  to see if there are any obvious outliers.
png("outputs/ClusterTree_Postcut.png", 1000, 1000, pointsize=20)
plot(sampleTree2, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5,
     cex.axis = 1.5, cex.main = 2)
dev.off()
```

## 5. Updating matrices for following analyses.  
!!! do this whether or not you cut samples !!!
```{r}
dim(dm_vst_t)
dim(treatment_data)

# Form a data frame analogous to expression data that will hold the clinical traits.
samples = rownames(dm_vst_t);# start new variable 'd7.Samples' calling the row names of the gene data (sample as 'Mouse' in trait data)
TreatRows = match(samples, treatment_data$sample.name); # match the names - calls the list number of 'd7.Samples' matching 'd7.Treatment.data$Sample.Name'
Traits = treatment_data[TreatRows, -1]; # insert TreatRows as the row numbers in 'treatment_data'
rownames(Traits) = treatment_data[TreatRows, 1]; # inserts the new TreatRows - matches sample IDs
all(rownames(Traits) == rownames(dm_vst_t))  # should be TRUE
dim(Traits)
```


## 6. Prepare trait data
```{r}
# COMBINE TREATMENT DATA
all_traits <-   Traits %>% dplyr::select('phase.treat') %>% #build matrix with all treatment groups
  dplyr::mutate(p1_control = as.factor(as.numeric(phase.treat == "1_control")))  %>%  # call occurrence of 'control' as 0s and 1s (factor)
  dplyr::mutate(p1_stress = as.factor(as.numeric(phase.treat == "1_stress")))  %>%  # call occurrence of 'stress' as 0s and 1s (factor)
  dplyr::mutate(p2_control = as.factor(as.numeric(phase.treat == "2_control")))  %>%  # call occurrence of 'control' as 0s and 1s (factor)
  dplyr::mutate(p2_stress = as.factor(as.numeric(phase.treat == "2_stress")))  %>%  # call occurrence of 'stress' as 0s and 1s (factor)
  dplyr::mutate(p2_naive = as.factor(as.numeric(phase.treat == "2_naive")))  %>%  # call occurrence of 'naive' as 0s and 1s (factor)
  dplyr::mutate(p2_recovery = as.factor(as.numeric(phase.treat == "2_recovery")))  %>%  # call occurrence of 'recovery' as 0s and 1s (factor)
  dplyr::select(-phase.treat)
all_traits
```


## 7. Cluster samples by trait
*make sure you are using the right sample tree w/ outlier samples removed!!*
```{r}
png("outputs/AllTraits_ClusterTree.png", 1000, 1000, pointsize=20)
traitColors = labels2colors(all_traits); # Convert traits to a color representation: white means low, red means high, grey means missing entry
plotDendroAndColors(sampleTree2, traitColors, # Plot the sample dendrogram and the colors underneath.
                    groupLabels = names(all_traits), 
                    main = "Sample dendrogram and trait heatmap")
dev.off()

# write some of the data thus far.... 
#save(dm_vst_t, Traits, file = "data/dataInput.RData")
#write.csv(dm_vst_t, "data/vstTransformed_WGCNAdata.csv") # write the vst transformed data 
```

## 8. Set softpower threshold
The next thing to do is determine the power or soft threshold to apply for the adjacency matrix. This is because we want to make a weighted matrix, where each gene has a weighted connection, not a hard threshold of either connected or not. From my understanding, by raising connection/adjacency to a power there in lies the weighted-ness of their connection.

I need to pick a power where scale free topology of the matrix is met. From some googling, this is similar to a scale free network. "A scale-free network can be constructed by progressively adding nodes to an existing network and introducing links to existing nodes with preferential attachment so that the probability of linking to a given node is proportional to the number of existing links that node has" So there is a higher chance of being connected to a node gene if it has more connections. That makes intuitive sense, which is why it has to reach this criterion to be passable for further analysis. <https://mathworld.wolfram.com/Scale-FreeNetwork.html>

"If the scale-free topology fit index fails to reach values above 0.8 for reasonable powers (less than 15 for unsigned or signed hybrid networks, and less than 30 for signed networks) and the mean connectivity remains relatively high (in the hundreds or above), chances are that the data exhibit a strong driver that makes a subset of the samples globally different from the rest. The difference causes high correlation among large groups of genes which invalidates the assumption of the scale-free topology approximation."<https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/faq.html>

"If the lack of scale-free topology fit turns out to be caused by an interesting biological variable that one does not want to remove (i.e., adjust the data for), the appropriate soft-thresholding power can be chosen based on the number of samples as in the table below. This table has been updated in December 2017 to make the resulting networks conservative." For a signed network with less than 15 samples, 18 is an appropriate power to use.

<https://peterlangfelder.com/2018/11/25/signed-or-unsigned-which-network-type-is-preferable/>
```{r}
dim(dm_vst_t) # again double check you have the correct data...
# Choose a set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to=20, by=2))
# Call the network topology analysis function
sft = pickSoftThreshold(dm_vst_t, powerVector = powers, verbose = 5) #...wait for this to finish (~20mins w/ full dataset)
#  pickSoftThreshold 
#  performs the analysis of network topology and aids the
#  user in choosing a proper soft-thresholding power.
#  The user chooses a set of candidate powers (the function provides suitable default values)
#  function returns a set of network indices that should be inspected

# png to output 
sizeGrWindow(9, 5) # set window size 
png("outputs/ScaleTopology_SoftThresh.png", 1000, 1000, pointsize=20)
        par(mfrow = c(1,2));
        cex1 = 0.9;
        
        plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2], # Scale-free topology fit index as a function of the soft-thresholding power
             xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
             main = paste("Scale independence"));
        text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
             labels=powers,cex=cex1,col="red");
        abline(h=0.8,col="red") # look at at cut off at power of 3 - this line corresponds to using an R^2 cut-off of h
        
        plot(sft$fitIndices[,1], sft$fitIndices[,5], # Mean connectivity as a function of the soft-thresholding power
             xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
             main = paste("Mean connectivity"))
        text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")
dev.off() # output 


# The left panel... shows the scale-free fit index (y-axis) 
# as a function of the soft-thresholding power (x-axis). 
# We choose the power 8, which is the lowest power for which the scale-free 
# topology index curve flattens out upon reaching a high value (in this case, roughly 0.90).
# The right panel.... displays the mean connectivity
# (degree, y-axis) as a function of the soft-thresholding power (x-axis).
```
These data look good based on the plots here. 

**I'll pick _8_ as my softpower threshold because this is where the Scale Free Topology R sq value is >=0.9 and the curve starts to flatten out.**


## 9. Stepwise module construction
#### Step 1 = create adjacency matrix 
https://peterlangfelder.com/2018/11/25/signed-or-unsigned-which-network-type-is-preferable/
https://www.rdocumentation.org/packages/WGCNA/10cpm/versions/1.69/topics/adjacency
https://ramellose.github.io/networktutorials/wgcna.html
https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/10cpm/TechnicalReports/signedTOM.pdf

```{r}
softPower = 8 # set your soft threshold based on the plots above 

# signed - must call the type, defaults to unsigned
adjacency_sign = adjacency(dm_vst_t, power = softPower, type="signed") # this takes time.. just wait...(~10mins for full dataset)
```

#### Step 2: Turn adjacency into topological overlap matrix
Calculation of the topological overlap matrix, (TOM)
and the corresponding dissimilarity, from a given adjacency matrix.

# signed matrix
```{r}
TOM_sign       = TOMsimilarity(adjacency_sign, TOMType="signed")  # this takes a long time.. just wait... (~2.5hrs for the full dataset)
dissTOM_sign   = 1-TOM_sign
```

#### Step 3:Call the hierarchical clustering function - plot the tree
```{r}
#Call the hierarchical clustering function
geneTree_sign   = hclust(as.dist(dissTOM_sign), method = "average");

# Plot the resulting clustering tree (dendrogram)
sizeGrWindow(12,9)
plot(geneTree_sign, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity - SIGNED",
     labels = FALSE, hang = 0.04)
```
#### Step 4: Set module size and 'cutreeDynamic' to create clusters 

```{r}
# We like large modules, so we set the minimum module size relatively high:
minModuleSize = 50; # set this for the subseqent call - WGCNA authors recommend diligence when calling module size to avoid too many/too few modules...
# Module identification using dynamic tree cut:
dynamicMods_sign = cutreeDynamic(dendro = geneTree_sign, distM = dissTOM_sign,
                                 deepSplit = 1, pamRespectsDendro = FALSE,
                                 minClusterSize = minModuleSize);
table(dynamicMods_sign) # view the number of genes per module 
```

#### Step 5: convert numeric network to colors and plot the dendrogram
```{r}
# Convert numeric labels into colors
dynamicColors_sign = labels2colors(dynamicMods_sign) # add colors to module labels (previously numbers)
table(dynamicColors_sign) # lets look at this table...

# Plot the dendrogram and colors underneath
sizeGrWindow(8,6)
plotDendroAndColors(geneTree_sign, dynamicColors_sign, "Dynamic Tree Cut - SIGNED",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05,
                    main = "Gene dendrogram and module colors 'SIGNED'")
```
#### Step 6: Calculate Eigengenes - view their connectivity based on 'MEDiss = 1-cor(MEs)'
```{r}
# Calculate eigengenes ========================================================== # 

# MEList = moduleEigengenes(dim1_mvg, colors = dynamicColors)
MEList = moduleEigengenes(dm_vst_t, colors = dynamicColors_sign)
MEs    = MEList$eigengenes # you can view MEs, condenses gene counts down to a single number for each sample representative of that expression pattern 

#remove module grey - option if you have a module grey (unassigned genes) - I didn't here
#MEs <- MEs[c(1,2,4,5)]

# Calculate dissimilarity of module eigengenes
MEDiss = 1-cor(MEs)
METree = hclust(as.dist(MEDiss), method = "average") # Cluster module eigengenes

# grouping the clusters from a cut-off
MEDissThres = 0.25

# Plot the result ================================================================ #
sizeGrWindow(7, 6)
png("outputs/ClusterEigengenes_cutoff.png", 1000, 1000, pointsize=20)
plot(METree, main = "Clustering of module eigengenes - SIGNED (dissimilarity calc = MEDiss = 1-cor(MEs))",
     xlab = "", sub = "")
abline(h=MEDissThres, col = "red")
dev.off()
```

#### Step 7: Specify the cut line for the dendrogram (module) - Calc MODULE EIGENGENES (mergeMEs)
Here, all of my modules are more 25% different from each other so I won't be merging any. Default cut height for merging modules is less than 25% dissimilarity. I could explore merging modules mint and forest (30% different) and modules iris and plum (27% different), but I don't find that this is necessary. More information on merging modules: https://fuzzyatelin.github.io/bioanth-stats/module-F21-Group1/module-F21-Group1.html#Module_Merging
```{r}
# MEDiss
# # merge modules whose distance is less than given cutHeight
# merge <- mergeCloseModules(dm_vst_t, dynamicColors_sign, cutHeight = .35)
# 
# # The merged module colors, assigning one color to each module
# mergedColors = merge$colors
# 
# # Eigengenes of the new merged modules
# mergedMEs = merge$newMEs
```


#### Step 8: Plot dendrogram with the cut line 'MEDissThres' ---- NOT USING HERE!
use 'mergedColors' we needed to merge related eigengene modules together 
```{r}
# sizeGrWindow(12, 9)
# png("outputs/ClusterDendrogram.png", 1000, 1000, pointsize=20)
# plotDendroAndColors(geneTree_sign, cbind(dynamicColors_sign, mergedColors),
#                     c("Original Modules", "Merged Modules"),
#                     dendroLabels = FALSE, hang = 0.03,
#                     addGuide = TRUE, guideHang = 0.05,
#                     main = "Gene dendrogram and module colors for original and merged modules")
# dev.off()
```

#### Step 9: Commit to mergedcolors as 'MEs' and 'moduleColors'
```{r}
# Rename to moduleColors
moduleColors = dynamicColors_sign

# Construct numerical labels corresponding to the colors
#colorOrder = c("grey", standardColors(50)); #I didn't have any genes that were not assigned so I won't keep grey here
colorOrder = c(standardColors(50))
moduleLabels = match(moduleColors, colorOrder); #add -1 if module grey present
#customColorOrder = c("goldenrod", "honeydew3", "palegreen", "seagreen", "thistle", "mediumpurple", "khaki2", "plum4") 
customColorOrder = c("goldenrod", "honeydew3", "palegreen", "seagreen", "thistle", "mediumpurple", "khaki2", "plum4") 
moduleColors.custom = customColorOrder[moduleLabels]

# Save module colors and labels for use in subsequent parts
save(MEs, moduleLabels, moduleColors.custom, file = "data/networkConstruction-stepByStep.RData")
# write csv - save the module eigengenes
write.csv(MEs, file = "data/WGCNA_ModuleEigengenes.csv")
```

Side quest - plot module eigengene dissimilarity tree using display names (for supplemental)
```{r}
# #pull display names as the column names for MEs data
# MEs_display_names <- MEs
# MEs_display_names
# colnames(MEs_display_names) <- c("gold",      "lemon",      "sage",     "mint",    "forest",   "lilac",         "iris",  "plum")
# MEs_display_names
# 
# # Calculate dissimilarity of module eigengenes
# MEDiss_display = 1-cor(MEs_display_names)
# METree_display = hclust(as.dist(MEDiss_display), method = "average") # Cluster module eigengenes
# 
# # grouping the clusters from a cut-off
# MEDissThres = 0.25
# 
# # Plot the result ================================================================ #
# sizeGrWindow(7, 6)
# png("outputs/ClusterEigengenes_cutoff_displayColors.png", 1000, 1000, pointsize=20)
# plot(METree_display, main = "Clustering of module eigengenes - SIGNED (dissimilarity calc = MEDiss = 1-cor(MEs))",
#      xlab = "", sub = "")
# abline(h=MEDissThres, col = "red")
# dev.off()
```

## 10. Prepare for module trait associations
Identify modules that are significantly associated with the measured  traits (here as treatment)

Since we already have a summary profile (eigengene) for each module,
we simply correlate eigengenes with external traits and look for the  significant associations:
```{r}
# Define numbers of genes and samples
nGenes = ncol(dm_vst_t); 
nSamples = nrow(dm_vst_t); 
# Recalculate MEs with color labels
MEs0 = moduleEigengenes(dm_vst_t, moduleColors.custom)$eigengenes
MEs = orderMEs(MEs0) # reorders the columns (colors/modules)
```

## 11. Module trait coorelation 
```{r}
# ALL TRAIT DATA
dim(Traits)  # 64  8
dim(MEs)  # 64  7

# treatment coorelation
all_traits_asnum    <- data.frame(lapply(all_traits, function(x) as.numeric(as.character(x))),
                                check.names=F, row.names = row.names(all_traits))
moduleTraitCor     = cor(MEs, all_traits_asnum, use = "p");
moduleTraitPvalue   = corPvalueStudent(moduleTraitCor, nSamples);
print("Module Correlation p-values")
moduleTraitPvalue
```

____*NAME KEY!!*____
"MEgoldenrod" = gold
"MEkhaki2" = lemon
"MEhoneydew3" = sage    
"MEpalegreen" = mint
"MEseagreen" = forest     
"MEthistle" = lilac      
"MEmediumpurple" = iris
"MEplum4" = plum



## 12. Write out modules to csv files (for GO enrichment)

First, we have to create a new matrix with the display names from the heatmaps that match up to the proper modules
```{r}
# change moduleColors.custom to new names (in the same order!!! - very important**)
#module_names = c("goldenrod", "honeydew3", "palegreen", "seagreen", "thistle", "mediumpurple", "khaki2", "plum4")  #original module name
display_names = c("gold",      "sage",      "mint",     "forest",    "lilac",   "iris",         "lemon",  "plum")   #module name for plot
moduleColors.display = display_names[moduleLabels]

# make data frame from dds object to pull from 
dds_modules = as.data.frame((dm_vst_t))
dim(dds_modules) 

# Prove that the display name is referencing the correct module  ***
mod_genes <- names(dds_modules)[moduleColors.custom=="goldenrod"] 
display_genes <- names(dds_modules)[moduleColors.display=="gold"]
length(mod_genes)==length(display_genes) # should be TRUE
mod_genes == display_genes # should all be TRUE
```


```{r}
# assign modules to variables
gold_genes <- names(dds_modules)[moduleColors.display=="gold"] 
lemon_genes <- names(dds_modules)[moduleColors.display=="lemon"] 
sage_genes <- names(dds_modules)[moduleColors.display=="sage"] 
mint_genes <- names(dds_modules)[moduleColors.display=="mint"] 
forest_genes <- names(dds_modules)[moduleColors.display=="forest"] 
lilac_genes <- names(dds_modules)[moduleColors.display=="lilac"] 
iris_genes <- names(dds_modules)[moduleColors.display=="iris"] 
plum_genes <- names(dds_modules)[moduleColors.display=="plum"] 

#save list of genes that comprise each module
write.csv(gold_genes, "module_lists/module_gold_genes.csv", quote=F, row.names = FALSE)
write.csv(lemon_genes, "module_lists/module_lemon_genes.csv", quote=F, row.names = FALSE)
write.csv(sage_genes, "module_lists/module_sage_genes.csv", quote=F, row.names = FALSE)
write.csv(mint_genes, "module_lists/module_mint_genes.csv", quote=F, row.names = FALSE)
write.csv(forest_genes, "module_lists/module_forest_genes.csv", quote=F, row.names = FALSE)
write.csv(lilac_genes, "module_lists/module_lilac_genes.csv", quote=F, row.names = FALSE)
write.csv(iris_genes, "module_lists/module_iris_genes.csv", quote=F, row.names = FALSE)
write.csv(plum_genes, "module_lists/module_plum_genes.csv", quote=F, row.names = FALSE)
```


Set color legend for heatmap 
```{r}
#rename modules for heatmap - ensure that they line up with names(MEs)
names(MEs)
#names(MEs) =     "MEgoldenrod","MEkhaki2", "MEhoneydew3", "MEpalegreen","MEseagreen", "MEthistle", "MEmediumpurple", "MEplum4"
color_legend <- c("gold",       "lemon",    "sage",         "mint",      "forest",     "lilac",      "iris",          "plum")
```



## 13. Heatmaps!
```{r}
sizeGrWindow(10,10)
all.matrix <- paste(signif(moduleTraitCor, 2), "\n(",
                      signif(moduleTraitPvalue, 1), ")", sep = "")

# Force it back into a matrix matching moduleTraitCor
dim(all.matrix) <- dim(moduleTraitCor)

# make function for heatmap colors
colfunc <- colorRampPalette(c("cornflowerblue", "grey100", "maroon"))

# set x-axis names - make sure they align with names(all_traits)
names(all_traits)
treatment_names = c("Phase I Control", "Phase I Stress", "Phase II Control", "Phase II Stress", "Phase II Naive", "Phase II Recovery")

#par(mar = c(8, 9.5, 5, 3))
png("heatmaps/WGCNA_heatmap.png", 800, 800, pointsize=16)

par(mar = c(8, 7, 5, 3),
    font = 2)

# textMatrix = NULL so the function draws NO text itself
labPos <- labeledHeatmap(Matrix = moduleTraitCor,
               xLabels = treatment_names,
               yLabels = names(MEs),
               #ySymbols = names(MEs),
               ySymbols = color_legend,
               colorLabels = TRUE,
               colors = colfunc(50), 
               textMatrix = NULL,
               setStdMargins = FALSE,
               cex.text = 1,
               zlim = c(-0.6,0.6),
               font.lab.x = 2, #bold x labels
               font.lab.y = 2, #bold y labels
               #main = paste("Module-trait relationships")
               )

# Separator between Phase I (columns 1–2) and Phase II (columns 3–6)
phase_boundary <- mean(labPos$xMid[c(2, 3)])

abline(
  v = phase_boundary,
  col = "black",
  lwd = 2,
  xpd = FALSE
)

# Draw every label exactly once, bolding significant cells
for (i in 1:nrow(moduleTraitCor)) {
  for (j in 1:ncol(moduleTraitCor)) {
    isSig <- moduleTraitPvalue[i, j] < 0.05
    text(x = labPos$xMid[j], y = labPos$yMid[i],
         labels = all.matrix[i, j],
         font = ifelse(isSig, 2, 1), cex = 1, adj = c(0.5, 0.5))
  }
}
dev.off()
```