---
title: "DESeq2_analysis"
author: "Megan Delatte-Guidry"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Differential gene expression analysis with DESeq2
DESeq2 estimates variance-mean dependence in count data and tests for differential expression base on a negative binomial distribution model.   

DESeq2 for Analyzing RNAseq data manual:
https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html

Input files:
- non-normalized count matrix
- metadata file

Note on non-normalized data from DESeq2 manual: "As input, the DESeq2 package expects count data as obtained, e.g., from RNA-seq or another high-throughput sequencing experiment, in the form of a matrix of integer values. The value in the i-th row and the j-th column of the matrix tells how many reads can be assigned to gene i in sample j. Analogously, for other types of assays, the rows of the matrix might correspond e.g. to binding regions (with ChIP-Seq) or peptide sequences (with quantitative mass spectrometry). We will list method for obtaining count matrices in sections below.

The values in the matrix should be un-normalized counts or estimated counts of sequencing reads (for single-end RNA-seq) or fragments (for paired-end RNA-seq). The RNA-seq workflow describes multiple techniques for preparing such count matrices. It is important to provide count matrices as input for DESeq2’s statistical model (Love, Huber, and Anders 2014) to hold, as only the count values allow assessing the measurement precision correctly. The DESeq2 model internally corrects for library size, so transformed or normalized values such as counts scaled by library size should not be used as input."


Load libraries
```{r, message=FALSE, warning=FALSE}
library(DESeq2)
library(data.table)
library(dplyr)
library(tidyr)
library(reshape2)
library(apeglm)
library(ggplot2)
library(vsn)
library(pheatmap)
library(RColorBrewer)
library(genefilter)
library(rsconnect)
library(gplots)
library(ashr)
library(limma)
```

## Load in data
```{r}
# load in metadata
sample_info <- read.csv("juve_cado_metadata_w_batches.csv", header=TRUE, sep=",", row.names=1)
head(sample_info)
tail(sample_info)
```

```{r}
# load in gene count matrix and set the row names header as "gene_id"
#gene_counts <- as.data.frame(read.csv("gene_count_matrix.csv", row.names="gene_id"))
gene_counts <- read.csv("gene_count_matrix.csv", row.names = 1, check.names = FALSE)
head(rownames(gene_counts))
head(gene_counts[,1:6])
```

## Check sample names
Manual: "It is absolutely critical that the columns of the count matrix and the rows of the column data (information about samples) are in the same order. DESeq2 will not make guesses as to which column of the count matrix belongs to which row of the column data, these must be provided to DESeq2 already in consistent order."  

The sample names in `sample_info` and `gene_counts` have to match up and be in the exact same order!
```{r}
# remove the "_F" in the column names of `gene_counts`
colnames(gene_counts) <- sub("_F", "", colnames(gene_counts))
head(gene_counts)

#check that the sample names are the same in both places
all(rownames(sample_info) %in% colnames(gene_counts))
#check the order of the samples
all(rownames(sample_info) == colnames(gene_counts))
```

## Subsetting data into PHASE I and PHASE II
```{r}
# subset metadata ('sample_info') first
sample_info <- tibble::rownames_to_column(sample_info, var = "sample.name")

p1_sample_info <- subset(sample_info, phase=="phase1")
p2_sample_info <- subset(sample_info, phase=="phase2")
dim(p1_sample_info)
dim(p2_sample_info)

#split count matrix based on names in each metadata variable
p1_gene_counts <- gene_counts[, p1_sample_info$sample.name]
dim(p1_gene_counts)
p2_gene_counts <- gene_counts[, p2_sample_info$sample.name]
dim(p2_gene_counts)
```

## Pre-filtering with genefilter
While pre-filtering of low count genes isn't necessary, it can be useful for 2 reasons:    
1. reduce size of the `dds` data object (thus increase speed of count modeling).  
2. improve visualizations by making them a little cleaner
```{r}
###filtering values for PoverA
#set filter values for PoverA (P percent of the samples have counts over A)
# here P = 0.04 and A = 5 (4% of samples would be 3 out of 72 samples)
filt <- filterfun(pOverA(0.04,5))

#create filter for the counts data
p1_tfil <- genefilter(p1_gene_counts, filt)
p2_tfil <- genefilter(p2_gene_counts, filt)

#identify transcripts to keep by count filter
p1_keep <- p1_gene_counts[p1_tfil,]
p2_keep <- p2_gene_counts[p2_tfil,]

#identify gene names in transcript list to keep
p1_gn.keep <- rownames(p1_keep)
p2_gn.keep <- rownames(p2_keep)

#data filtered in PoverA, P percent of the samples have counts over A
p1_filt_gene_counts <- as.data.frame(p1_gene_counts[which(rownames(p1_gene_counts) %in% p1_gn.keep),])
write.csv(p1_filt_gene_counts, file="p1_filtered_gene_count_matrix.csv")
head(p1_filt_gene_counts,10)

p2_filt_gene_counts <- as.data.frame(p2_gene_counts[which(rownames(p2_gene_counts) %in% p2_gn.keep),])
write.csv(p2_filt_gene_counts, file="p2_filtered_gene_count_matrix.csv")
head(p2_filt_gene_counts,10)
```
#> nrow(gene_counts)
#[1] 97257
#> nrow(filt_gene_counts)
#[1] 70981



## PHASE I - DESeq2 analysis & visualization

Building DESeq2 matrix
```{r}
# assign the batch numbers as factors not integers
p1_sample_info$batch <- factor(p1_sample_info$batch, levels = c("1", "2"))

#build deseq matrix (dm#)
dm1 <- DESeqDataSetFromMatrix(countData = p1_filt_gene_counts,
                             colData = p1_sample_info,
                             design = ~ batch + treatment)
dm1$treatment <- factor(dm1$treatment, levels = c("control", "stress"))
levels(dm1$treatment)
dm1
```

Running DESeq2
```{r}
dds1 <- DESeq(dm1) #takes a couple minutes to run
res1<-results(dds1) 
head(res1)
summary(res1)
resultsNames(dds1)
```

Investigating the models 
To investigate the models, logfold shrinkage was viewed with MAplots. 
```{r}
dds1_shrink <- lfcShrink(dds1, contrast =c("treatment","control", "stress"), type="ashr")
## using 'ashr' for LFC shrinkage. If used in published research, please cite:
##     Stephens, M. (2016) False discovery rates: a new deal. Biostatistics, 18:2.
##     https://doi.org/10.1093/biostatistics/kxw041
```

**plotMA** 
"In DESeq2, the function plotMA shows the log2 fold changes attributable to a given variable over the mean of normalized counts for all the samples in the DESeqDataSet.   
Points will be colored red if the adjusted p value is less than 0.1. Points which fall out of the window are plotted as open triangles pointing either up or down." Using the ashr shrinkage option allows for specifying the contrasts in the results object."
```{r}
DESeq2::plotMA(dds1_shrink, ylim=c(-5,5))
```

Setting the alpha value and creating results sets
Here I'll set the alpha value as a more conservative 0.05 (compared to the default 0.1). 
Note: It is important to specific the correct order in the contrasts, with the name of the numerator level for the fold change, and the name of the denominator level for the fold change for the comparison.
```{r}
summary(res1)
sum(res1$padj!=0, na.rm=TRUE) #35715 DEGs total
sum(res1$padj<0.1, na.rm=TRUE) #532 pass the filter with 0.1 cutoff
sum(res1$padj<0.05, na.rm=TRUE) #304 pass the filter with 0.05 cutoff
```

```{r}
#filtering based on alpha = 0.05 (which is the cutoff value for the adjusted p-value) where to be included the DEG's have to have an adj. p-value less than alpha
resultsNames(dds1)

res1_comp <- results(dds1, alpha=0.05, contrast=c("treatment","control", "stress"))
```

```{r}
summary(res1_comp)   

#then lets save these results set for later use (pretty sure this is what I'll use for the GO term analysis)
#write.csv(res1_comp,"outputs/phase1_res_stress.csv")
```

Next, I'll subset each results set to only be significantly DEGs and order then by the adjusted p value (small to large)
```{r}
sig_res1_comp <- subset(res1_comp, padj<0.05) #identify signficant pvalues with 5%FDR
summary(sig_res1_comp)
#order by padj
ordered_sig_res1_comp <-sig_res1_comp[ order(sig_res1_comp$padj ), ] 

#these ordered lists were written to files for potential later use - like with gProfiler
write.csv(ordered_sig_res1_comp,"outputs/sig_phase1_res_stress.csv")
```

#### Visualization
For most of the analysis/visualization I wanted to compare the treatments accross all significant DEGs between all the treatments. So I combined the significant subsets and removed duplicates by using the original matrix, as well as regularize log transformed the list for the PCA and heatmap visualization.

```{r}
#combine significant DEGs for all treatment contrasts
combine1 <- rbind(sig_res1_comp)
combine_list1 <- dm1[which(rownames(dm1) %in% rownames(combine1)),] #combined list from the original matrix so dupicates are removed 
vsd1 <- varianceStabilizingTransformation(combine_list1, blind=FALSE)
```

limma removeBatchEffect()
```{r}
mat1 <- assay(vsd1)
mm1 <- model.matrix(~treatment, colData(vsd1))
mat1 <- limma::removeBatchEffect(mat1, batch=vsd1$batch, design=mm1)
assay(vsd1) <- mat1
```

Principle Components Analysis
```{r}
# Quick PCA plot to start
plotPCA(vsd1, intgroup=c("treatment")) # same PCA as below but less pretty, use this for knowing the PC loadings for the labels in the below PCA

# prettier pca
pcaData <- plotPCA(vsd1, intgroup=c("treatment"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
percentVar
ggplot(pcaData, aes(PC1, PC2, color=treatment, label=name)) +
  geom_point(aes(fill=treatment), pch = 21, size = 7, color= "white") +
  #geom_text() +
  coord_fixed() + ylab("PC2: 10% Variance Explained") +
  xlab("PC1:37% Variance Explained") +
  #theme_linedraw() +
  scale_fill_manual(values = c(control="#A29BAD", stress ="#D29F13")) +
  theme_minimal(base_size = 20) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.title=element_text(face="bold", size=24, color="white"),
    axis.text=element_text(size=24, color="white"), 
    axis.text.x = element_text(size = 20),
    panel.border = element_rect(color = "white", fill = NA, linewidth = 1),
    panel.background = element_rect(fill = alpha("white", 0.1))
    )


ggsave("pca_plots/phase1_pca_5batch_colorbytreatment_blk_ann.png", last_plot(), width=7, height=7)
```

Heatmap of rlog transformed difference in expression compared to the average across all samples, using adjusted p value of 0.00001 or less. Many thanks to Hollie Putnam for the code of how to make this heatmap.
```{r}
sig.num <- sum(combine1$padj <0.00001, na.rm=T) #set the p value cut off very low here, 54 of 298
sum(combine1$padj <0.00001, na.rm=T)
sum(combine1$padj!=0)
topVarGenes <- head(order(rowVars(assay(vsd1)),decreasing=TRUE),sig.num) #sort by decreasing sig
mat <- assay(vsd1)[ topVarGenes, ] #make an expression object
mat <- mat - rowMeans(mat) #difference in expression compared to average across all samples
#col.order <- c("CASE_J03", "CASE_J09", "CASE_J12", "CASE_J13", "CA_J06",   "CA_J08",   "CA_J11",   "CA_J18",   "CON_J02",  "CON_J05" , "CON_J10" , "SE_J01" ,  "SE_J04",   "SE_J07")
#mat <- mat[,col.order]
df1 <- as.data.frame(colData(vsd1)[c("treatment")]) #make dataframe for column naming 


colfunc <- colorRampPalette(c("deepskyblue", "white", "violetred3")) #make function for the color gradient 
ann_colors <- list( treatment= c(control="#A29BAD", stress ="#D29F13"))
breakss <- c(-2, -1.9, -1.8, -1.7, -1.6, -1.5, -1.4, -1.3, -1.2, -1.1, -1, -.9, -.8, -.7, -.6, -.5, -.4, -.3, -.2, -.1, 0, .1, .2, .3, .4, .5, .6, .7, .8, .9, 1, 1.1, 1.2, 1.3, 1.4, 1.5, 1.6, 1.7, 1.8, 1.9, 2) #this looks very extra but this is how the colors in the heatmap were broken up 


png("phase1_heatmap.png", units= "in", width=20, height=20, res = 600)
p <- pheatmap(mat, annotation_col=df1, annotation_colors=ann_colors, annotation_names_col = F, clustering_method = "average", 
         clustering_distance_rows="euclidean", show_rownames =FALSE, cluster_cols=F,
         show_colnames =F, breaks= breakss, color = colfunc(40), annotation_legend = F) 
dev.off()
```

## PHASE II - DESeq2 analysis & visualization

Building DESeq2 matrix
```{r}
# assign the batch numbers as factors not integers
p2_sample_info$batch <- factor(p2_sample_info$batch, levels = c("2","3", "4", "5"))

#build deseq matrix (dm#)
dm2 <- DESeqDataSetFromMatrix(countData = p2_filt_gene_counts,
                             colData = p2_sample_info,
                             design = ~ batch + treatment)
dm2$treatment <- factor(dm2$treatment, levels = c("control", "naive","recovery", "stress"))
levels(dm2$treatment)
dm2
```


Running DESeq2
```{r}
dds2 <- DESeq(dm2) #takes a couple minutes to run
res2<-results(dds2) 
head(res2)
summary(res2)
resultsNames(dds2)
```

Investigating the models 
To investigate the models, logfold shrinkage was viewed with MAplots. 
```{r}
dds_1_cs <- lfcShrink(dds, contrast =c("phase.treat","1_control", "1_stress"), type="ashr")
dds_1s2s <- lfcShrink(dds, contrast =c("phase.treat","1_stress", "2_stress"), type="ashr")
dds_2s2n <- lfcShrink(dds, contrast =c("phase.treat","2_stress", "2_naive"), type="ashr")
## using 'ashr' for LFC shrinkage. If used in published research, please cite:
##     Stephens, M. (2016) False discovery rates: a new deal. Biostatistics, 18:2.
##     https://doi.org/10.1093/biostatistics/kxw041
```

**plotMA** 
"In DESeq2, the function plotMA shows the log2 fold changes attributable to a given variable over the mean of normalized counts for all the samples in the DESeqDataSet.   
Points will be colored red if the adjusted p value is less than 0.1. Points which fall out of the window are plotted as open triangles pointing either up or down." Using the ashr shrinkage option allows for specifying the contrasts in the results object."
```{r}
DESeq2::plotMA(dds_2s2n, ylim=c(-5,5))
```

Setting the alpha value and creating results sets

Here I'll set the alpha value as a more conservative 0.05 (compared to the default 0.1). 
Note: It is important to specific the correct order in the contrasts, with the name of the numerator level for the fold change, and the name of the denominator level for the fold change for the comparison.
```{r}
summary(res2)
sum(res2$padj!=0, na.rm=TRUE) #46599 DEGs total
sum(res2$padj<0.1, na.rm=TRUE) #1081 pass the filter with 0.1 cutoff
sum(res2$padj<0.05, na.rm=TRUE) #760 pass the filter with 0.05 cutoff
```

```{r}
#filtering based on alpha = 0.05 (which is the cutoff value for the adjusted p-value) where to be included the DEG's have to have an adj. p-value less than alpha
resultsNames(dds2)

res_stress <- results(dds2, alpha=0.05, contrast=c("treatment","control", "stress"))
res_naive <- results(dds2, alpha=0.05, contrast=c("treatment","control", "naive")) 
res_recovery <- results(dds2, alpha=0.05, contrast=c("treatment","control", "recovery"))
```

```{r}
summary(res_stress)   
summary(res_naive) 
summary(res_recovery) 


#then lets save these results set for later use (pretty sure this is what I'll use for the GO term analysis)
#write.table(res_stress,"outputs/res_stress.txt",quote=FALSE,col.names=TRUE,row.names=TRUE,sep="\t")
#write.table(res_naive,"outputs/res_naive.txt",quote=FALSE,col.names=TRUE,row.names=TRUE,sep="\t")
#write.table(res_recovery,"outputs/res_recovery.txt",quote=FALSE,col.names=TRUE,row.names=TRUE,sep="\t")
```

Next, I'll subset each results set to only be significantly DEGs and order then by the adjusted p value (small to large)
```{r}
sig_res_stress <- subset(res_stress, padj<0.05) #identify signficant pvalues with 5%FDR
sig_res_naive <- subset(res_naive, padj<0.05)
sig_res_recovery <- subset(res_recovery, padj<0.05)



#order by padj
ordered_sig_res_stress <-sig_res_stress[ order(sig_res_stress$padj ), ] 
ordered_sig_res_naive <-sig_res_naive[ order(sig_res_naive$padj ), ]
ordered_sig_res_recovery <-sig_res_recovery[ order(sig_res_recovery$padj ), ]

#these ordered lists were written to files for potential later use - like with gProfiler
#write.table(ordered_sig_res_stress,"outputs/ordered_sig_res_stress.txt",quote=FALSE,col.names=TRUE,row.names=TRUE,sep="\t")
#write.table(ordered_sig_res_naive,"outputs/ordered_sig_res_naive.txt",quote=FALSE,col.names=TRUE,row.names=TRUE,sep="\t")
#write.table(ordered_sig_res_recovery,"outputs/ordered_sig_res_recovery.txt",quote=FALSE,col.names=TRUE,row.names=TRUE,sep="\t")

```


#### Visualization
For most of the analysis/visualization I wanted to compare the treatments accross all significant DEGs between all the treatments. So I combined the significant subsets and removed duplicates by using the original matrix, as well as regularize log transformed the list for the PCA and heatmap visualization.

```{r}
#combine significant DEGs for all treatment contrasts
combine2 <- rbind(sig_res_stress, sig_res_naive, sig_res_recovery)
combine_list2 <- dm2[which(rownames(dm2) %in% rownames(combine2)),] #combined list from the original matrix so dupicates are removed 
#combine_rlog <- rlog(combine_list, blind=FALSE) #setting blind to FALSE takes into account levels and contrasts
vsd2 <- varianceStabilizingTransformation(combine_list2, blind=FALSE)

dim(combine_list2)
```

limma removeBatchEffect()
```{r}
mat <- assay(vsd2)
mm <- model.matrix(~treatment, colData(vsd2))
mat <- limma::removeBatchEffect(mat, batch=vsd2$batch, design=mm)
assay(vsd2) <- mat
```

```{r}
# Quick PCA plot to start
plotPCA(vsd2, intgroup=c("treatment")) # same PCA as below but less pretty, use this for knowing the PC loadings for the labels in the below PCA

# prettier pca
pcaData <- plotPCA(vsd2, intgroup=c("treatment"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
percentVar
ggplot(pcaData, aes(PC1, PC2, color=treatment, label=name)) +
  geom_point(aes(fill=treatment), pch = 21, size = 7, color= "white") +
  #geom_text() +
  coord_fixed() + ylab("PC2: 8% Variance Explained") +
  xlab("PC1:34% Variance Explained") +
  #theme_linedraw() +
  scale_fill_manual(values = c(control="#A29BAD", stress ="#D29F13", naive= "#6BA84C", recovery= "#65A6E2")) +
  theme_minimal(base_size = 20) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.title=element_text(face="bold", size=24, color="white"),
    axis.text=element_text(size=24, color="white"), 
    axis.text.x = element_text(size = 20),
    panel.border = element_rect(color = "white", fill = NA, linewidth = 1),
    panel.background = element_rect(fill = alpha("white", 0.1))
    )

ggsave("phase2_pca_5batch_colorbytreatment_blk_ann.png", last_plot(), width=7, height=7)
```

plotting other PCs
```{r}
# Get matrix of transformed counts
vst_mat <- assay(vsd)

# Transpose for PCA: samples in rows
pca <- prcomp(t(vst_mat))

# Build a data frame with metadata and all PCs
pcaData <- as.data.frame(pca$x)
pcaData$treatment <- colData(vsd)$treatment
pcaData$name <- colnames(vsd)

# Get percent variance for axis labels
percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100

# Plot PC3 vs PC4
ggplot(pcaData, aes(PC2, PC3, color=treatment, label=name)) +
  geom_point(size=3) +
  coord_fixed() +
  ylab(paste0("PC3: ", round(percentVar[4]), "% Variance Explained")) +
  xlab(paste0("PC2: ", round(percentVar[3]), "% Variance Explained")) +
  theme_linedraw()
```

Heatmap of rlog transformed difference in expression compared to the average across all samples, using adjusted p value of 0.00001 or less. Many thanks to Hollie Putnam for the code of how to make this heatmap.
```{r}
sig.num <- sum(combine2$padj <0.00001, na.rm=T) #set the p value cut off very low here, 1174 of 9060
sum(combine2$padj <0.00001, na.rm=T)
sum(combine2$padj!=0)
topVarGenes <- head(order(rowVars(assay(vsd2)),decreasing=TRUE),sig.num) #sort by decreasing sig
mat <- assay(vsd2)[ topVarGenes, ] #make an expression object
mat <- mat - rowMeans(mat) #difference in expression compared to average across all samples
#col.order <- c("CASE_J03", "CASE_J09", "CASE_J12", "CASE_J13", "CA_J06",   "CA_J08",   "CA_J11",   "CA_J18",   "CON_J02",  "CON_J05" , "CON_J10" , "SE_J01" ,  "SE_J04",   "SE_J07")
#mat <- mat[,col.order]
df2 <- as.data.frame(colData(vsd2)[c("treatment")]) #make dataframe for column naming 

colfunc <- colorRampPalette(c("deepskyblue", "white", "violetred3")) #make function for the color gradient 
ann_colors <- list( treatment= c(control="#A29BAD", stress ="#D29F13", naive= "#6BA84C", recovery= "#65A6E2"))
breakss <- c(-2, -1.9, -1.8, -1.7, -1.6, -1.5, -1.4, -1.3, -1.2, -1.1, -1, -.9, -.8, -.7, -.6, -.5, -.4, -.3, -.2, -.1, 0, .1, .2, .3, .4, .5, .6, .7, .8, .9, 1, 1.1, 1.2, 1.3, 1.4, 1.5, 1.6, 1.7, 1.8, 1.9, 2) #this looks very extra but this is how the colors in the heatmap were broken up 

png("phase2_heatmap.png", units= "in", width=20, height=20, res = 600)
pheatmap(mat, annotation_col=df2, annotation_colors=ann_colors, annotation_names_col = F, clustering_method = "average", 
         clustering_distance_rows="euclidean", show_rownames =FALSE, cluster_cols=F,
         show_colnames =F, breaks= breakss, color = colfunc(40), annotation_legend = F) 
dev.off()
```


Let's make a Venn diagram of the significant DEGs that overlap across treatments and count how many overlap between the treatments. This is using modified code from [here](https://www.biostars.org/p/288028/).
```{r}
sig_res_naive_genes <- row.names(sig_res_naive) #just gets the names of the genes, which is all that is needed to compare
sig_res_recovery_genes <- row.names(sig_res_recovery)
sig_res_stress_genes <- row.names(sig_res_stress)


comb2 <- c(sig_res_naive_genes, sig_res_recovery_genes, sig_res_stress_genes) #combine the list

comb2_list <- dm2[which(rownames(dm2) %in% comb2),] #gets rid of any duplicates 

comb2_row <- row.names(comb2_list) #make just the names of the genes again

# Comparing each individual list with the combined list
sig_res_naive_genes.3 <- comb2_row %in% sig_res_naive_genes

sig_res_recovery_genes.3 <- comb2_row %in% sig_res_recovery_genes

sig_res_stress_genes.3 <- comb2_row %in% sig_res_stress_genes 


## COMPARING ALL 3 TREATMENTS!
# Generating venn counts to plot venn diagram 
counts3 <- cbind(sig_res_naive_genes.3, sig_res_recovery_genes.3, sig_res_stress_genes.3)
ven3 <- vennCounts(counts3)
png("phase2_venn-diagram-all-treatments.png", 1000, 1000)
vennDiagram(ven3, cex=1, names = c("naive", "recovery", "stress"), circle.col= c("#6BA84C", "#65A6E2", "#D29F13"))
dev.off()
```


Venn diagram comparing Phase 1 and 2 stress!
```{r}
#phase 1 - get just gene names
sig_res1_genes <- row.names(sig_res1_comp)
#phase 2 - get just gene names
sig_res_stress_genes <- row.names(sig_res_stress)

comb <- c(sig_res1_genes, sig_res_stress_genes) #combine the list

comb_list <- dm2[which(rownames(dm2) %in% comb),] #gets rid of any duplicates 

comb_row <- row.names(comb_list) #make just the names of the genes again


# Comparing each individual list with the combined list
sig_res_stress_genes_comp <- comb_row %in% sig_res_stress_genes
sig_res1_genes_comp <- comb_row %in% sig_res1_genes

# Generating venn counts to plot venn diagram 
counts <- cbind(sig_res1_genes_comp, sig_res_stress_genes_comp)
ven <- vennCounts(counts)
png("phase2_venn-diagram-p1&2.png", 1000, 1000)
vennDiagram(ven, cex=1, names = c("phase1", "phase2"))
dev.off()
```








