---
title: "GO Enrichment of DEGs for Phase I and II"
author: "Megan Delatte-Guidry"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(topGO)
library(stringr)
library(dplyr)
library(tidyr)
```

### Steps

1. Data preparation: List of genes identifiers, gene scores, list of differentially expressed genes or a criteria for selecting genes based on their scores, as well as gene-to-GO annotations are all collected and stored in a single R object.   
(referenced [Amy Zyck's TopGO script](https://github.com/MarineEvoEcoLab/ROD_CADO/blob/main/Outlier_Exploration/Outlier_GO.Rmd) for the gene2go table)

2. Running the enrichment tests: Using the object created in the first step the user can perform enrichment analysis using any feasible mixture of statistical tests and methods that deal with the GO topology.

3. Analysis of the results: The results obtained in the second step are analysed using summary functions and visualization tools.
_________________________________________________________________________________


### gene2go table from annotation information
PULLING gene2go_full.tab FROM THE WGCNA DIRECTORY (../wgcna/gene2go_full.tab)

gaf from NCBI with the GO terms for genes in the eastern oyster genome. I'm going to take this file and extract the Gene IDs and associated GO terms.

In terminal:
```{bash}
#wget https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/002/022/765/GCF_002022765.2_C_virginica-3.0/GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf.gz
#gunzip GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf.gz
#mv GCF_002022765.2_C_virginica-3.0_gene_ontology.gaf gene_ontology.gaf
```

Now I can make a new .tab file with just the gene ID and GO IDs (3rd and 5th columns) renamed "gene_id" and "GO_Annotation". The gene_name column will have one unique observation per row and all associated GO terms will be saved in the same row, with values separated by ;
```{bash}
#awk 'NR>9 {print $3 "\t" $5}' gene_ontology.gaf | 
#sed '1i gene_id\tGO_Annotation' | 
#sort | awk '
#NR==1 {print; next}
#$1!=prev {
#if (NR>2) print prev "\t" goids;
#prev=$1;
#goids=$2;
#next
#}
#{
#goids=goids "; " $2
#}
#END {print prev "\t" goids}
#' > gene2go_full.tab
```


### Gene IDs for DEGs are in the following files:
outputs/phase1_directional_expression.csv
outputs/phase2_naive_directional_expression.csv
outputs/phase2_recovery_directional_expression.csv
outputs/phase2_stress_directional_expression.csv

Make separate files with gene IDs and GO IDs for UP & DOWN regulated genes
```{r}
# set dataset 
phasetreat <- "p2_recovery"

# load data (DEG tables filtered for Padj<0.05 and |LFC|>1)
deg <- read.csv(paste0("outputs/",phasetreat,"_DEGs_sig.05.csv"), header = TRUE, sep = ",")
deg <- deg[,(2:6)] #remove vector column bc I goofed on DEG file write out
#View(deg)
# separate UP and DOWN reg genes into two data sets
deg_up <- subset(deg, diffexpressed=="UP")
deg_down <- subset(deg, diffexpressed=="DOWN")

# load go annotation file to assign go terms to color module
go_terms <- read.table("../wgcna/gene2go_full.tab", header = TRUE, sep="\t")
head(go_terms)

# UP - reformat gene_id column to be just LOC#, add in go annotation, remove 'MSTRG' rows and NA rows
deg_up$gene_id <- str_extract(deg_up$GeneID, "[^|]+$")
dim(deg_up)
deg_up_wGO <- merge(deg_up, go_terms, by = "gene_id", all.x = TRUE)
dim(deg_up_wGO)
#clean up
deg_up_wGO <- na.omit(deg_up_wGO)
deg_up_wGO <- deg_up_wGO[, c("gene_id", "GO_Annotation")]
dim(deg_up_wGO)

# DOWN - reformat gene_id column to be just LOC#, add in go annotation, remove 'MSTRG' rows and NA rows
deg_down$gene_id <- str_extract(deg_down$GeneID, "[^|]+$")
dim(deg_down)
deg_down_wGO <- merge(deg_down, go_terms, by = "gene_id", all.x = TRUE)
dim(deg_down_wGO)
#clean up
deg_down_wGO <- na.omit(deg_down_wGO)
deg_down_wGO <- deg_down_wGO[, c("gene_id", "GO_Annotation")]
dim(deg_down_wGO)

# write out tables
write.table(deg_up_wGO, paste0("topGO/data/",phasetreat,"_UP_genes_GO_header.tab"), sep = "\t", row.names=FALSE)
write.table(deg_down_wGO, paste0("topGO/data/",phasetreat,"_DOWN_genes_GO_header.tab"), sep = "\t", row.names=FALSE)
```

Read in gene2go information 
```{r}
all_gene2go<-read.delim("../wgcna/gene2go_full.tab", sep="\t")
UP_gene2go<-read.delim(paste0("topGO/data/",phasetreat,"_UP_genes_GO_header.tab"), sep="\t")
DOWN_gene2go<-read.delim(paste0("topGO/data/",phasetreat,"_DOWN_genes_GO_header.tab"), sep="\t")
View(UP_gene2go)
View(DOWN_gene2go)
```

Make list of genes for input to TopGO
```{r}
UP_genes <- as.character(unique(UP_gene2go$gene_id))
DOWN_genes <- as.character(unique(DOWN_gene2go$gene_id))

all_genes <- as.character(unique(all_gene2go$gene_id))

UP_GeneList <- factor(as.integer(all_genes %in% UP_genes))
names(UP_GeneList) <- all_genes

DOWN_GeneList <- factor(as.integer(all_genes %in% DOWN_genes))
names(DOWN_GeneList) <- all_genes
```

The following code will perform GO enrichment using the weighted Fisher's exact test  

Read in gene-to-go-mappings
```{r}
gene2go_topgo <-readMappings("../wgcna/gene2go_full.tab", IDsep=";", sep="\t")
```

Set function to select genes of interest (ie those that have pvalue < 0.05)
```{r}
topDiffGenes <- function(allScore) {
return(allScore < 0.05)}
```

### Biological Processes
```{r}
#Create `topGOdata` object, which is required for topGO analysis
GO_UP_BP <-new("topGOdata", ontology="BP", gene2GO=gene2go_topgo, allGenes=UP_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
GO_DOWN_BP <-new("topGOdata", ontology="BP", gene2GO=gene2go_topgo, allGenes=DOWN_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)


#Run GO enrichment test 
GO_UP_BP_FE <- runTest(GO_UP_BP, algorithm="weight01", statistic="fisher")
GO_DOWN_BP_FE <- runTest(GO_DOWN_BP, algorithm="weight01", statistic="fisher")


#Generate results table 
GO_UP_BP_RT <- GenTable(GO_UP_BP, Fisher = GO_UP_BP_FE, orderBy = "Fisher",  topNodes = 50, numChar = 100)
GO_DOWN_BP_RT <- GenTable(GO_DOWN_BP, Fisher = GO_DOWN_BP_FE, orderBy = "Fisher",  topNodes = 50, numChar = 100)
#Only taking the top 100 GO terms 

#Filter by significant results
GO_UP_BP_RT$Fisher<-as.numeric(GO_UP_BP_RT$Fisher)
GO_UP_BP_RT_sig<-GO_UP_BP_RT[GO_UP_BP_RT$Fisher<0.05,] #sig UP
GO_DOWN_BP_RT$Fisher<-as.numeric(GO_DOWN_BP_RT$Fisher)
GO_DOWN_BP_RT_sig<-GO_DOWN_BP_RT[GO_DOWN_BP_RT$Fisher<0.05,] #sig DOWN
```

Merge `GO_UP_BP_RT_sig` with GO and gene info. 
```{r}
# Separate GO terms 
UP_gene2go <- UP_gene2go %>%
  separate_rows(GO_Annotation, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
UP_gene2go$GO_Annotation <- trimws(UP_gene2go$GO_Annotation)
GO_UP_BP_RT_sig$GO.ID <- trimws(GO_UP_BP_RT_sig$GO.ID)

# Join the datasets based on GO term
GO_UP_BP_RT_sig_gene <- UP_gene2go %>%
  left_join(GO_UP_BP_RT_sig, by = c("GO_Annotation" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_UP_BP_RT_sig_gene$ontology <- "Biological Processes"
```

Merge `GO_DOWN_BP_RT_sig` with GO and gene info. 
```{r}
# Separate GO terms 
DOWN_gene2go <- DOWN_gene2go %>%
  separate_rows(GO_Annotation, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
DOWN_gene2go$GO_Annotation <- trimws(DOWN_gene2go$GO_Annotation)
GO_DOWN_BP_RT_sig$GO.ID <- trimws(GO_DOWN_BP_RT_sig$GO.ID)

# Join the datasets based on GO term
GO_DOWN_BP_RT_sig_gene <- DOWN_gene2go %>%
  left_join(GO_DOWN_BP_RT_sig, by = c("GO_Annotation" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_DOWN_BP_RT_sig_gene$ontology <- "Biological Processes"
```

### Cellular Components
```{r}
#Create `topGOdata` object, which is required for topGO analysis
GO_UP_CC <-new("topGOdata", ontology="CC", gene2GO=gene2go_topgo, allGenes=UP_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
GO_DOWN_CC <-new("topGOdata", ontology="CC", gene2GO=gene2go_topgo, allGenes=DOWN_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)

#Run GO enrichment test 
GO_UP_CC_FE <- runTest(GO_UP_CC, algorithm="weight01", statistic="fisher")
GO_DOWN_CC_FE <- runTest(GO_DOWN_CC, algorithm="weight01", statistic="fisher")

#Generate results table 
GO_UP_CC_RT <- GenTable(GO_UP_CC, Fisher = GO_UP_CC_FE, orderBy = "Fisher",  topNodes = 50, numChar = 100)
GO_DOWN_CC_RT <- GenTable(GO_DOWN_CC, Fisher = GO_DOWN_CC_FE, orderBy = "Fisher",  topNodes = 50, numChar = 100)
#Only taking the top 100 GO terms 

#Filter by significant results
GO_UP_CC_RT$Fisher<-as.numeric(GO_UP_CC_RT$Fisher)
GO_UP_CC_RT_sig<-GO_UP_CC_RT[GO_UP_CC_RT$Fisher<0.05,] #sig UP
GO_DOWN_CC_RT$Fisher<-as.numeric(GO_DOWN_CC_RT$Fisher)
GO_DOWN_CC_RT_sig<-GO_DOWN_CC_RT[GO_DOWN_CC_RT$Fisher<0.05,] #sig DOWN
```

Merge `GO_UP_CC_RT_sig` with GO and gene info. 
```{r}
# Separate GO terms 
UP_gene2go <- UP_gene2go %>%
  separate_rows(GO_Annotation, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
UP_gene2go$GO_Annotation <- trimws(UP_gene2go$GO_Annotation)
GO_UP_CC_RT_sig$GO.ID <- trimws(GO_UP_CC_RT_sig$GO.ID)

# Join the datasets based on GO term
GO_UP_CC_RT_sig_gene <- UP_gene2go %>%
  left_join(GO_UP_CC_RT_sig, by = c("GO_Annotation" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_UP_CC_RT_sig_gene$ontology <- "Cellular Components"
```

Merge `GO_DOWN_CC_RT_sig` with GO and gene info. 
```{r}
# Separate GO terms 
DOWN_gene2go <- DOWN_gene2go %>%
  separate_rows(GO_Annotation, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
DOWN_gene2go$GO_Annotation <- trimws(DOWN_gene2go$GO_Annotation)
GO_DOWN_CC_RT_sig$GO.ID <- trimws(GO_DOWN_CC_RT_sig$GO.ID)

# Join the datasets based on GO term
GO_DOWN_CC_RT_sig_gene <- DOWN_gene2go %>%
  left_join(GO_DOWN_CC_RT_sig, by = c("GO_Annotation" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_DOWN_CC_RT_sig_gene$ontology <- "Cellular Components"
```


### Molecular Functions
```{r}
#Create `topGOdata` object, which is required for topGO analysis
GO_UP_MF <-new("topGOdata", ontology="MF", gene2GO=gene2go_topgo, allGenes=UP_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)
GO_DOWN_MF <-new("topGOdata", ontology="MF", gene2GO=gene2go_topgo, allGenes=DOWN_GeneList, annot = annFUN.gene2GO, geneSel=topDiffGenes)

#Run GO enrichment test 
GO_UP_MF_FE <- runTest(GO_UP_MF, algorithm="weight01", statistic="fisher")
GO_DOWN_MF_FE <- runTest(GO_DOWN_MF, algorithm="weight01", statistic="fisher")

#Generate results table 
GO_UP_MF_RT <- GenTable(GO_UP_MF, Fisher = GO_UP_MF_FE, orderBy = "Fisher",  topNodes = 50, numChar = 100)
GO_DOWN_MF_RT <- GenTable(GO_DOWN_MF, Fisher = GO_DOWN_MF_FE, orderBy = "Fisher",  topNodes = 50, numChar = 100)
#Only taking the top 100 GO terms 

#Filter by significant results
GO_UP_MF_RT$Fisher<-as.numeric(GO_UP_MF_RT$Fisher)
GO_UP_MF_RT_sig<-GO_UP_MF_RT[GO_UP_MF_RT$Fisher<0.05,] #sig UP
GO_DOWN_MF_RT$Fisher<-as.numeric(GO_DOWN_MF_RT$Fisher)
GO_DOWN_MF_RT_sig<-GO_DOWN_MF_RT[GO_DOWN_MF_RT$Fisher<0.05,] #sig DOWN
```

Merge `GO_UP_MF_RT_sig` with GO and gene info. 
```{r}
# Separate GO terms 
UP_gene2go <- UP_gene2go %>%
  separate_rows(GO_Annotation, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
UP_gene2go$GO_Annotation <- trimws(UP_gene2go$GO_Annotation)
GO_UP_MF_RT_sig$GO.ID <- trimws(GO_UP_MF_RT_sig$GO.ID)

# Join the datasets based on GO term
GO_UP_MF_RT_sig_gene <- UP_gene2go %>%
  left_join(GO_UP_MF_RT_sig, by = c("GO_Annotation" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_UP_MF_RT_sig_gene$ontology <- "Molecular Functions"
```

Merge `GO_DOWN_MF_RT_sig` with GO and gene info. 
```{r}
# Separate GO terms 
DOWN_gene2go <- DOWN_gene2go %>%
  separate_rows(GO_Annotation, sep = ";")

# Ensure GO terms in both datasets are formatted similarly (trim whitespaces)
DOWN_gene2go$GO_Annotation <- trimws(DOWN_gene2go$GO_Annotation)
GO_DOWN_MF_RT_sig$GO.ID <- trimws(GO_DOWN_MF_RT_sig$GO.ID)

# Join the datasets based on GO term
GO_DOWN_MF_RT_sig_gene <- DOWN_gene2go %>%
  left_join(GO_DOWN_MF_RT_sig, by = c("GO_Annotation" = "GO.ID")) %>%
  na.omit()

# Add ontology column 
GO_DOWN_MF_RT_sig_gene$ontology <- "Molecular Functions"
```

## Join ontologies
Bind `GO_UP_BP_RT_sig_gene`, `GO_UP_CC_RT_sig_gene`, and `GO_UP_MF_RT_sig_gene` to make a df that has significantly enriched GO terms for all ontologies 
```{r}
GO_UP_sig_genes <- rbind(GO_UP_BP_RT_sig_gene, GO_UP_CC_RT_sig_gene, GO_UP_MF_RT_sig_gene)
GO_DOWN_sig_genes <- rbind(GO_DOWN_BP_RT_sig_gene, GO_DOWN_CC_RT_sig_gene, GO_DOWN_MF_RT_sig_gene)
```

Merge `pi_bed_filt` with `GO_UP_sig_genes` and calculate proportion of significant genes
```{r}
merged_up_genes <- GO_UP_sig_genes %>%
  mutate(prop.sig.genes = Significant/Annotated)

# saving unique instances of gene/GO pairs
merged_up_genes <- unique(merged_up_genes)

merged_down_genes <- GO_DOWN_sig_genes %>%
  mutate(prop.sig.genes = Significant/Annotated)

# saving unique instances of gene/GO pairs
merged_down_genes <- unique(merged_down_genes)
```

```{r}
# Write to csv 
write.csv(merged_up_genes, paste0("topGO/data/enrichment_tables/",phasetreat,"_UP_genes_GO_enrichment.csv"))
write.csv(merged_down_genes, paste0("topGO/data/enrichment_tables/",phasetreat,"_DOWN_genes_GO_enrichment.csv"))
```