---
title: "GO_MWU notes and file prep"
author: "Megan Delatte-Guidry"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Rank-based Gene Ontology analysis of gene expression data with [`GO_MWU`](https://github.com/z0on/GO_MWU) 

## Results summary 
Kind of worked with these file inputs, though I will investigate TopGO, gprofiler, and other options for GO enrichment.    
See below for the numbers of GO terms that passed the 10% FDR threshold for both phases.

Phase 1: (1 MF; 1 BP; 0 CC)
Phase 2: stress (0 MF; 1 BP; 4 CC)
         recovery (6 MF; 12 BP; 7 CC)
         naive ( 0 MF; 7 BP; 0 CC)

**Notes:**
- There was nothing terribly exciting identified with this analysis.  
- I cannot figure out how to print out those sinlge GO terms (bc of the way the gomwu function builds the results table). However for those with >1 GO term.   
- plots saved in ./gomwu_plots


## Short Guide (pulled from package github):
1. Put all this into the same directory:
  * scripts: GO_MWU.R, gomwu_a.pl, gomwu_b.pl, gomwu.functions.R
  * GO hierarchy file (go.obo, http://www.geneontology.org/GO.downloads.ontology.shtml)
  * table of GO annotations for your sequences: two-column (gene id - GO terms), tab-delimited, one line per gene, multiple GO terms separated by semicolon. If you have multiple lines per gene, use nrify_GOtable.pl to merge them. Do NOT include genes without GO annotations.
  * table of measure of interest for your sequences: two columns of comma-separated values: gene id, continuous measure of change such as                 log(fold-change). To perform standard GO enrichment analysis based on Fisher's exact test, use binary measure (1 or 0, i.e., either sgnificant or not). To analyze modules derived from WGCNA, specify 0 for genes not included in the module and the kME value (number between 0 and 1, module membership score) for genes included in the module.

It is important to have the latter two tables representing the whole genome (or transcriptome) - at least the portion that was measured - rather than some select group of genes since the test relies on comparing the behavior of individual GO categories to the whole.

## Input file formatting details 
[Yaamini's notbook post](https://yaaminiv.github.io/DML-Analysis-Part39/) was super helpful!

**goAnnotation input file!**
- reformat large GO annotation table (from Sam) 
- pull only GeneID & Annotation_GO_ID columns
- blank -> unknown
- csv -> .tab

**table of significance measures**
- generate full DEG table (for each phase) w/o removing n.s. data with de_analysis/DESeq2_analysis.Rmd
- reformat GeneID column of DEG table
- calculate signed negative log p-value and output table with gene id and signed neg log p-value


### Let's start with the goAnnotation file!

This should be good to run across the board. 

copy big GO Annotation file to this directory
```{bash}
#
cp ../de_analysis/Cviginiva_GOterms.csv .
mv Cviginiva_GOterms.csv table_formatting/

# next steps (in R): 
# - pull only GeneID and Annotation_GO_ID columns
# - change blanks to "unknown"
# - save file as .tab w/o quotes and row names
```

formatting
```{r}
#load packages
library(dplyr)

# read in file
go_table <- read.csv("table_formatting/Cviginiva_GOterms.csv",
                     header = TRUE, stringsAsFactors = FALSE)
#pull out only GeneID column and Annotation_GO_ID column
go_table2 <- go_table %>% select(GeneID, Annotation_GO_ID)
#clean up table - reassign blank or NA GO annotations as "unknown"
go_table2_clean <- go_table2 %>%
  mutate(Annotation_GO_ID = as.character(Annotation_GO_ID)) %>%
  mutate(Annotation_GO_ID = ifelse(
    is.na(Annotation_GO_ID) | Annotation_GO_ID == "",
    "unknown",
    Annotation_GO_ID
  ))
#write out table with out vector numbers and quotations
write.table(go_table2_clean, "table_formatting/GO_annotation_Cvirg.tab",
            sep = "\t", row.names = FALSE, quote = FALSE)


#check for duplicates in GO table
sum(duplicated(go_table2_clean$GeneID))

dim(go_table2_clean)
```

Lots of repeated LOC GeneIDs so, as per the GO MWU documentation, I'll run the nrify_GOtable.pl script to deal with duplicates before proceeding with the GO_MWU.R script. 
```
perl nrify_GOtable.pl table_formatting/GO_annotation_Cvirg.tab > dup_GO_annotation_Cvirg.tab
```

Check repeats
```{r}
dup_GO <- read.table("dup_GO_annotation_Cvirg.tab")
#check for duplicates in GO table
sum(duplicated(dup_GO$GeneID))
dim(dup_GO)
```


### Moving on to the table of significance measures!

Copy over full DEG table from phase 1 & 2(including n.s. data)
```{bash}
cp ../de_analysis/outputs/phase1_res_stress.csv table_formatting/
cp ../de_analysis/outputs/phase2_res_stress.csv table_formatting/
cp ../de_analysis/outputs/phase2_res_recovery.csv table_formatting/
cp ../de_analysis/outputs/phase2_res_naive.csv table_formatting/

# next steps (in R):
# - pull out only GeneID column and p-value column??
# - reformat GeneID columns (to match up w/ the LOC# in the Go table)
# - calculate the signed negative log p-value
# - output table with two columns: GeneID and signed neg. p-value
```

Here, I calculated the log transformed p-values for each GeneID signed (+/-) base on whether that gene was up (+) or down (-) regulated
```{r}
library(stringr)

deg <- read.csv("table_formatting/phase2_res_naive.csv",
                     header = TRUE, stringsAsFactors = FALSE)
# make a new column called "GeneID" in the 'deg' df that is only the gene ID
deg$GeneID <- str_extract(deg$X, "[^|]+$")

# 1. calulate the negative log of the p-values
log_deg <- -log(deg$pvalue)
# 2. add negative log pvalues to dataframe & rename column
log_p1 <- data.frame(matrix(unlist(log_deg), nrow=length(log_deg), byrow=T)) 
colnames(log_p1) <- c("logpvalue") 
# 3. add column to deg data frame
deg <- cbind(deg, log_p1) 
# 4. take abs value of the log2fold change
av_lfc <- abs(deg$log2FoldChange)
# 5. make that list into a data frame & rename column
av_p1 <- data.frame(matrix(unlist(av_lfc), nrow=length(av_lfc), byrow=T)) 
colnames(av_p1) <- c("absolute_value_log_2change") 
# 6. add column to deg data frame
deg <- cbind(deg, av_p1) 
# 7. multiply the -log pvalue by the log2fold change (this will make downreg genes negative!)
deg$mult_logpvalue <- deg$logpvalue * deg$log2FoldChange 
# 8. divide the mulitpled -logpvalue by the absolute value of the log2fold change
deg[, "neglogpvalue"] <- deg[, "mult_logpvalue"] / deg[, "absolute_value_log_2change"] 

#pull out only GeneID column and neg log p-value column
deg2 <- deg %>% select(GeneID, neglogpvalue)
# omit rows with NAs and "MSTRG"
deg3 <- na.omit(deg2)
deg4 <- deg3 %>%
  filter(!str_detect(GeneID, "MSTRG"))

#check
dim(deg2)
dim(deg3)
dim(deg4)

# write out file as a .csv for input to GO_MWU script
write.csv(deg4, "phase2_NAIVE_logtrans_signed_pvalues.csv", row.names = FALSE, quote= FALSE)

```


